### 2017-04-13 05_analysis_scripts/summary_table1.Rmd
updating summary table scripts

---
title: "summary_table1"
author: "dk"
date: "4/13/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r morph_table}
# This code calculates a mean, standard deviation, and N for each measured variable by region & lake type

# read the summary file
# need to figure out how to read file directly from GitHub
nes_all <- read.csv("../nes_data.csv", header=TRUE, na.strings="NA", stringsAsFactors = FALSE)

# add a column to specify region
nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "NORTHEASTERN"
nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "EASTERN"
nes_all[nes_all$state %in% c("ARKANSAS", "IOWA", "KANSAS", "LOUISIANA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "OKLAHOMA", "SOUTH DAKOTA", "TEXAS"), "region"] <- "CENTRAL"
nes_all[nes_all$state %in% c("ARIZONA", "CALIFORNIA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON", "WYOMING"), "region"] <- "WESTERN"
# make REGION a factor
nes_all$region <- as.factor(nes_all$region)
# reorder region levels so final table reads from west to east 
nes_all$region <- factor(nes_all$region, levels = c("WESTERN", "CENTRAL", "NORTHEASTERN", "EASTERN"))

# reshape data to [n]es_[a]ll [l]ong format
library(reshape2)

# nes_all[,9:34] <- apply(nes_all[,9:34], 2, as.numeric)

nal <- melt(nes_all, 
            id.vars = c("state", "name", "region", "lake_type"),
            measure.vars = c((names(nes_all)[c(9:34)])))
nal$value <- as.numeric(nal$value)

# mean(
#   as.numeric(nal[nal$variable == "p_surface_area_loading" & 
#            nal$region == "EASTERN" & 
#            nal$lake_type == "NATURAL", "value"])
#   , na.rm = TRUE)

# calculate mean, standard dev, and N for each variable by region and lake type
library(plyr)
summary_nes = ddply(nal, .(variable, region, lake_type), summarize,
                mean = mean(value, na.rm=TRUE),
                sd = sd(value, na.rm=TRUE),
                N = length(value[!is.na(value)]))

# any(!is.na(summary_nes$mean))

# now how to report numbers? let's convert them to character strings & round means & sd's to 2 decimal places for now
summary_nes$mean <- as.character(round(summary_nes$mean, digits=2))
summary_nes$sd <- as.character(round(summary_nes$sd, digits=2))
summary_nes$N <- as.character(summary_nes$N)

# add parentheses around N
summary_nes$N <- paste("(", summary_nes$N, ")", sep="")

# now create a column with the 3 values & the +- symbol (using paste?)
summary_nes$stat <- paste(summary_nes$mean, "\u00b1", summary_nes$sd, summary_nes$N, sep=" ")
# Didn't need to do this so far, but might need to execute the following if +- symbol doesn't work
Encoding(summary_nes$stat) <- "UTF-8"

# now name classify variables into morphometry, physicochemical, & loading variables
summary_nes[summary_nes$variable %in% c(levels(summary_nes$variable)[1:5]), "variable_type"] <- "MORPHOMETRY"
summary_nes[summary_nes$variable %in% c(levels(summary_nes$variable)[6:12]), "variable_type"] <- "PHYSIOCHEMICAL"
summary_nes[summary_nes$variable %in% c(levels(summary_nes$variable)[13:28]), "variable_type"] <- "LOADING"

# let's make three different dataframes for each of the variable types (each will be a separate table)
morph <- summary_nes[summary_nes$variable_type=="MORPHOMETRY", ]
physchem <- summary_nes[summary_nes$variable_type=="PHYSIOCHEMICAL", ]
load <- summary_nes[summary_nes$variable_type=="LOADING", ]

# let's reshape these
morph_wide <- dcast(morph, variable + lake_type ~ region, value.var="stat")
physchem_wide <- dcast(physchem, variable + lake_type ~ region, value.var="stat")  
load_wide <- dcast(load, variable + lake_type ~ region, value.var="stat")

# let's make the tables
# install.packages("devtools")
# devtools::install_github("rstudio/rmarkdown")
# For dev version
# devtools::install_github("haozhu233/kableExtra")

# load libraries
library(knitr)
library(kableExtra)
library(magrittr)

kable(morph_wide, format = "latex", booktabs = T) %>%
  kable_styling()
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2017-04-13 05_analysis_scripts/nes_summarystats.R

@@ -1,18 +1,18 @@
 # This code calculates a mean, standard deviation, and N for each measured variable by region & lake type
 
 # read the summary file
-# need to figure out how to read file directly from GitHub
+# make sure you open the Project in R
 nes_all <- read.csv("nes_data.csv", header=TRUE, na.strings="NA", stringsAsFactors = FALSE)
 
 # add a column to specify region
-nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "NORTHEASTERN"
-nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "EASTERN"
-nes_all[nes_all$state %in% c("ARKANSAS", "IOWA", "KANSAS", "LOUISIANA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "OKLAHOMA", "SOUTH DAKOTA", "TEXAS"), "region"] <- "CENTRAL"
-nes_all[nes_all$state %in% c("ARIZONA", "CALIFORNIA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON", "WYOMING"), "region"] <- "WESTERN"
+nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "Northeastern"
[…]

### 2017-04-14 05_analysis_scripts/MapData_Final.R
Failed attempt to added new axis labels

@@ -22,15 +22,15 @@ library(tmap)
 lakes.dat <- read.csv("./nes_data.csv")
 
 # Remove any rows with NA coordinates
-lakes.dat <- subset(lakes.dat, !is.na(Lat))
-lakes.dat <- subset(lakes.dat, !is.na(Long))
-which(is.na(lakes.dat$Lat) == TRUE) # This should now be zero
-which(is.na(lakes.dat$Lat) == TRUE) # This should now be zero
+lakes.dat <- subset(lakes.dat, !is.na(lat))
+lakes.dat <- subset(lakes.dat, !is.na(long))
+which(is.na(lakes.dat$lat) == TRUE) # This should now be zero
+which(is.na(lakes.dat$long) == TRUE) # This should now be zero
 
 # Make the longitudes negative if they aren't already
 for(i in 1:nrow(lakes.dat)){
-  if(lakes.dat$Long[i] > 0){
-    lakes.dat$Long[i] <- -lakes.dat$Long[i]
+  if(lakes.dat$long[i] > 0){
+    lakes.dat$long[i] <- -lakes.dat$Long[i]
   } 
 }
 
@@ -77,8 +77,8 @@ plot(us.nation)
 
 ########  IDW  ##########################################################
 # Create new columns for lat and long
-lakes.dat$x <- lakes.dat$Long  # define x & y as longitude and latitude - not really necessary...could just do with Long Lat in the next line
-lakes.dat$y <- lakes.dat$Lat 
+lakes.dat$x <- lakes.dat$long  # define x & y as longitude and latitude - not really necessary...could just do with Long Lat in the next line
+lakes.dat$y <- lakes.dat$lat 
 
 # Designate x and y columns as coordinates
 coordinates(lakes.dat) = ~x + y
@@ -109,11 +109,20 @@ plot(idw.ras)
 us.cond <- mask(idw.ras, us.unproj.nation)
 plot(us.cond)
 
+
+plot(us.cond, xaxt="n", yaxt = "n")
+axis(1, xaxp=c(10, 200, 19), las=2)
+axis(1, at = seq(10, 200, by = 10), las=2)
+
+
+
+
 # Reproject the raster in the us.atlas projection
 us.cond <- projectRaster(us.cond, crs = us.atlas.proj, over = T)
-plot(us.cond, main = "Conductivity")
+plot(us.cond, main = "Conductivity", ymin = y.range[1], ymax = y.range[2])
[…]

### 2017-04-20 README.Rmd
link to data in README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# National Eutrophication Survey

This respository contains a digitally transcribed copy ([nes_data.csv](https://github.com/ReproducibleQM/NES/blob/master/nes_data.csv)) of the National Eutrophication Survey (NES) dataset. The original data can be found in the `archival_pdfs` folder or by searching: https://www.epa.gov/nscep

## Locations

```{r eval=FALSE, echo=FALSE}
library(magick)
points <- image_read("06_images/points.jpeg")
image_write(image_trim(points), "06_images/points_trim.jpeg")
```

![](06_images/points_trim.jpeg)

## Data

| variable name | description              |units|
|---------------|--------------------------|-----|
|pdf|pdf identifier (474 - 477)| integer|
|pagenum | page number of the pdf|integer|
|storet_code     | identifier which links measurement to coordinate locations|character|
|state                | state where the water body resides |character|
|name               | name of the water body|character|
|county              | county where the water body resides |character|
|lake_type         | natural or impoundment |character|
|drainage_area | the total drainage area | square kilometers |
|surface_area   | the area of the water surface|sq km|
|mean_depth    | the volume of the water body divided by the surface area in square meters| cubic meters|
|total_inflow      | the mean of the inflows of all tributaries and the immediate drainage | cubic meters per second|
|retention_time | a mean value determined by dividing the lake volume, in cubic meters, by the mean annual outflow in cubic meters per unit cof time| years or days|
|retention_time_units | the units of time for each retention entry|years or days|
|alkalinity | |milligrams per liter|
|conductivity | |microohms|
|secchi | |meters|
|tp | total phosphorus| milligrams per liter|
|po4 | orthophosphate| milligrams per liter|
|tin | total inorganic nitrogen| milligrams per liter|
|tn | total nitrogen| milligrams per liter|
|p_pnt_source_muni | municipal point source phosphorus loading| kilograms per year|
|p_pnt_source_industrial | industrial point source phosphorus loading| kilograms per year|
|p_pnt_source_septic | septic point source phosphorus loading| kilograms per year|
|p_nonpnt_source | nonpoint source phosphorus loading| kilograms per year|
|p_total | total phosphorus loading| kilograms per year|
|n_pnt_source_muni | municipal point source nitrogen loading| kilograms per year|
|n_pnt_source_industrial | industrial point source nitrogen loading| kilograms per year|
|n_pnt_source_septic | septic point source nitrogen loading| kilograms per year|
|n_nonpnt_source | nonpoint source nitrogen loading| kilograms per year|
|n_total | total nitrogen loading| kilograms per year|
|p_total_out | total phosphorus outlet load| kilograms per year|
|p_percent_retention | percent phosphorus retention| percent|
|p_surface_area_loading | phosphorus surface area loading| grams per square meter per year|
|n_total_out | total nitrogen outlet load| kilograms per year|
|n_percent_retention | percent nitrogen retention|percent|
|n_surface_area_loading | nitrogen surface area loading| grams per square meter per year|
|lat | |decimal degrees|
|long | |decimal degrees|

## References

Stachelek, J., Ford, C., Kincaid, D., King, K., Miller, H., Nagelkirk, R. (_in prep_) The National Eutrophication Survey: lake characteristics and historical nutrient concentrations.

Stachelek, J. (2017). nesR: Scrape Data from National Eutrophication Survey archival PDFs. R package version 0.1. (https://github.com/jsta/nesR) (https://dx.doi.org/10.5281/zenodo.400258)

### 2017-04-20 05_analysis_scripts/MapData_Final.R
Changed color ramp and log transformed data in plots

@@ -15,6 +15,7 @@ library(ggplot2)
 library(gstat)
 library(raster)
 library(tmap)
+library(RColorBrewer)
 
 
 
@@ -73,7 +74,8 @@ us.nation <- spTransform(us.nation, us.atlas.proj)
 plot(us.48)
 plot(us.nation)
 
-
+# Define a color ramp for the maps
+my.palette <- brewer.pal(n = 9, name = "YlOrRd")
 
 ########  IDW  ##########################################################
 # Create new columns for lat and long
@@ -115,11 +117,9 @@ axis(1, xaxp=c(10, 200, 19), las=2)
 axis(1, at = seq(10, 200, by = 10), las=2)
 
 
-
-
 # Reproject the raster in the us.atlas projection
 us.cond <- projectRaster(us.cond, crs = us.atlas.proj, over = T)
-plot(us.cond, main = "Conductivity", ymin = y.range[1], ymax = y.range[2])
+plot(log(us.cond), main = "Conductivity", col = terrain.colors(255))  # rev(heat.colors(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
@@ -146,17 +146,17 @@ plot(us.alk)
 
 # Reproject the raster in the us.atlas projection
 us.alk <- projectRaster(us.alk, crs = us.atlas.proj, over = T)
-plot(us.alk, main = "Alkalinity")
+plot(log(us.alk), main = "Alkalinity", col = terrain.colors(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
 
 ############  SECCHI  ############################################################
 # Make a new df with no NAs for conductivity
-lakes.sec <- lakes.dat[complete.cases(lakes.dat$sechhi),]
+lakes.sec <- lakes.dat[complete.cases(lakes.dat$secchi),]
 
 # Apply the IDW
-idw <- idw(formula = sechhi ~ 1, locations = lakes.sec, newdata = grd)  # apply idw model for the data
+idw <- idw(formula = secchi ~ 1, locations = lakes.sec, newdata = grd)  # apply idw model for the data
 idw.ras <- raster(idw)
 
 # Rasterize and crop the IDW (might not be needed)
@@ -169,7 +169,7 @@ plot(us.sec)
 
 # Reproject the raster in the us.atlas projection
[…]

### 2017-04-20 05_analysis_scripts/merge_storet.R
rm old merging code

@@ -1,29 +1,0 @@
-#' @examples \dontrun{
-#' merge_storet("storet.csv", "476/res_review.csv")
-#' }
-
-merge_storet <- function(storet_path, data_path){
-  dt     <- read.csv(data_path, stringsAsFactors = FALSE)
-  zero_pad <- function(x){
-    if(!is.na(x)){
-      if(nchar(x) == 3){paste0("0", x)
-        }else{x}
-    }else{x}
-  }
-  dt$storet_code <- sapply(dt$storet_code, zero_pad)
-  
-  storet <- read.csv(storet_path, stringsAsFactors = FALSE)
-  storet$Long <- storet$Long * -1
-  
-  res <- merge(dt, storet, by.x = "storet_code", by.y = "Storet")
-  
-  browser()
-  library(sp)
-  library(maps)
-    coordinates(res) <- ~Long + Lat
-  map("state")
-  plot(res, add = TRUE)
-  res[which.min(coordinates(res)[,1]),]@data
-  
-  res
-}

### 2017-04-20 05_analysis_scripts/MapData_Final.R
Added new maps with red color ramp

@@ -77,6 +77,9 @@ plot(us.nation)
 # Define a color ramp for the maps
 my.palette <- brewer.pal(n = 9, name = "YlOrRd")
 
+
+pal <- colorRampPalette(c("white", "brown"))
+
 ########  IDW  ##########################################################
 # Create new columns for lat and long
 lakes.dat$x <- lakes.dat$long  # define x & y as longitude and latitude - not really necessary...could just do with Long Lat in the next line
@@ -119,7 +122,7 @@ axis(1, at = seq(10, 200, by = 10), las=2)
 
 # Reproject the raster in the us.atlas projection
 us.cond <- projectRaster(us.cond, crs = us.atlas.proj, over = T)
-plot(log(us.cond), main = "Conductivity", col = terrain.colors(255))  # rev(heat.colors(255))
+plot(log(us.cond), main = "Conductivity", col = pal(255))   #terrain.colors(255))  # rev(heat.colors(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
@@ -146,7 +149,7 @@ plot(us.alk)
 
 # Reproject the raster in the us.atlas projection
 us.alk <- projectRaster(us.alk, crs = us.atlas.proj, over = T)
-plot(log(us.alk), main = "Alkalinity", col = terrain.colors(255))
+plot(log(us.alk), main = "Alkalinity", col = pal(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
@@ -169,7 +172,7 @@ plot(us.sec)
 
 # Reproject the raster in the us.atlas projection
 us.sec <- projectRaster(us.sec, crs = us.atlas.proj, over = T)
-plot(log(us.sec), main = "Secchi", col = terrain.colors(255))
+plot(log(us.sec), main = "Secchi", col = pal(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
@@ -192,7 +195,7 @@ plot(us.p.total)
 
 # Reproject the raster in the us.atlas projection
 us.p.total <- projectRaster(us.p.total, crs = us.atlas.proj, over = T)
-plot(log(us.p.total), main = "P_Total", col = terrain.colors(255))
+plot(log(us.p.total), main = "P_Total", col = pal(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
[…]

### 2017-04-20 05_analysis_scripts/nes_summarystats.R
Added table for physchem section
Added a table for the physchem section. Working on last table (load) -
can't get the table to stop running off the page since it's so large.
I'll post it if I figure it out.

@@ -93,15 +93,22 @@ levels(summary_nes$lake_type)[levels(summary_nes$lake_type)=="IMPOUNDMENT"] <- "
 levels(summary_nes$lake_type)[levels(summary_nes$lake_type)=="NATURAL"] <- "Natural"
 
 # let's make three different dataframes for each of the variable types (each will be a separate table)
-morph <- summary_nes[summary_nes$variable_type=="MORPHOMETRY", ]
-physchem <- summary_nes[summary_nes$variable_type=="PHYSIOCHEMICAL", ]
-load <- summary_nes[summary_nes$variable_type=="LOADING", ]
+# morph <- summary_nes[summary_nes$variable_type=="MORPHOMETRY", ]
+# physchem <- summary_nes[summary_nes$variable_type=="PHYSIOCHEMICAL", ]
+# load <- summary_nes[summary_nes$variable_type=="LOADING", ]
+#Getting weird rows in dataframes created above, changed to use subset function
+
+morphdf <- subset(summary_nes, summary_nes$variable_type == "MORPHOMETRY")
+physchemdf <- subset(summary_nes, summary_nes$variable_type == "PHYSIOCHEMICAL")
+loaddf <- subset(summary_nes, summary_nes$variable_type == "LOADING")
+
+
 
 # THIS NEXT STEP BREAKS IT :(
 # let's reshape these
-morph_wide <- dcast(morph, variable + lake_type ~ region, value.var="stat")
-physchem_wide <- dcast(physchem, variable + lake_type ~ region, value.var="stat")  
-load_wide <- dcast(load, variable + lake_type ~ region, value.var="stat")
+morph_wide <- dcast(morphdf, variable + lake_type ~ region, value.var="stat")
+physchem_wide <- dcast(physchemdf, variable + lake_type ~ region, value.var="stat")  
+load_wide <- dcast(loaddf, variable + lake_type ~ region, value.var="stat")
 
 # let's make the tables
 # install.packages("devtools")
@@ -129,6 +136,15 @@ sink_text <- function(dt, fname){
 
 sink_text((kable(morph_wide, format = "latex", booktabs = T) %>%
   kable_styling()), "07_tables/morph_table.tex")
+
+sink_text((kable(physchem_wide, format = "latex", booktabs = T) %>%
+             kable_styling()), "07_tables/physchem_table.tex")
+
+#Cant get table to not run off page
[…]

### 2017-04-20 05_analysis_scripts/MapData_Final.R
Updated mapping code

@@ -74,6 +74,7 @@ us.nation <- spTransform(us.nation, us.atlas.proj)
 plot(us.48)
 plot(us.nation)
 
+
 # Define a color ramp for the maps
 my.palette <- brewer.pal(n = 9, name = "YlOrRd")
 
@@ -125,6 +126,9 @@ us.cond <- projectRaster(us.cond, crs = us.atlas.proj, over = T)
 plot(log(us.cond), main = "Conductivity", col = pal(255))   #terrain.colors(255))  # rev(heat.colors(255))
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
+
+
+
 
 
 # plot(us.unproj.nation, axes = TRUE, border = "white") # this will get the right axes

### 2017-04-26 README.Rmd
fill-in missing fields

@@
 
+<!-- Do not edit this table without first updating METADATA.md -->
+
 | variable name | description              |units|
@@
 |retention_time_units | the units of time for each retention entry|years or days|
-|alkalinity | |milligrams per liter|
-|conductivity | |microohms|
-|secchi | |meters|
+|alkalinity | alkalinity|milligrams per liter|
+|conductivity | conductivity|microohms|
+|secchi | secchi|meters|
 |tp | total phosphorus| milligrams per liter|
@@
 |n_surface_area_loading | nitrogen surface area loading| grams per square meter per year|
-|lat | |decimal degrees|
-|long | |decimal degrees|
+|lat | latitude|decimal degrees|
+|long | longitude|decimal degrees|
 

### 2017-04-27 05_analysis_scripts/nes_summarystats.R
rework tables without lake_type #11

@@ -1,114 +1,132 @@
-# This code calculates a mean, standard deviation, and N for each measured variable by region & lake type
+# This code calculates a mean, standard deviation, and N for each measured
+# variable by region & lake type ##############################################
 
-# read the summary file
-# make sure you open the Project in R
 nes_all <- read.csv("nes_data.csv", header=TRUE, na.strings="NA", stringsAsFactors = FALSE)
 
-# add a column to specify region
 nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "Northeastern"
 nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "Eastern"
 nes_all[nes_all$state %in% c("ARKANSAS", "IOWA", "KANSAS", "LOUISIANA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "OKLAHOMA", "SOUTH DAKOTA", "TEXAS"), "region"] <- "Central"
 nes_all[nes_all$state %in% c("ARIZONA", "CALIFORNIA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON", "WYOMING"), "region"] <- "Western"
-# make REGION a factor
 nes_all$region <- as.factor(nes_all$region)
-# reorder region levels so final table reads from west to east 
 nes_all$region <- factor(nes_all$region, levels = c("Western", "Central", "Northeastern", "Eastern"))
 
 # reshape data to [n]es_[a]ll [l]ong format
 library(reshape2)
 
-# nes_all[,9:34] <- apply(nes_all[,9:34], 2, as.numeric) 
-
 nal <- melt(nes_all, 
-            id.vars = c("state", "name", "region", "lake_type"),
+            id.vars = c("state", "name", "region"),
[…]

### 2017-05-16 05_analysis_scripts/PDF_Merge.R
rm row.names

@@ -60,4 +60,4 @@ df_final <- df_final[,c(
 
 # Export Dataframe as CSV -------------------------------------------------
 
-write.csv(df_final, file = "nes_data.csv")
+write.csv(df_final, file = "nes_data.csv", row.names = FALSE)

### 2017-05-16 README.Rmd
cite nesRdata, point to raw csv

@@
 
-This respository contains a digitally transcribed copy ([nes_data.csv](https://github.com/ReproducibleQM/NES/blob/master/nes_data.csv)) of the National Eutrophication Survey (NES) dataset.
+This respository contains a digitally transcribed copy ([nes_data.csv](https://github.com/ReproducibleQM/NES/raw/master/nes_data.csv)) of the National Eutrophication Survey (NES) dataset.
 The original data can be found in the `archival_pdfs` folder or by searching: https://www.epa.gov/nscep
@@
 Stachelek, J. (2017). nesR: Scrape Data from National Eutrophication Survey archival PDFs. R package version 0.1. (https://github.com/jsta/nesR) (https://dx.doi.org/10.5281/zenodo.400258)
+
+Stachelek, J. (2017). nesRdata: National Eutrophication Survey Data Package. R package version 0.1. https://github.com/jsta/nesRdata

### 2017-05-17 05_analysis_scripts/maps.R
Updated map script. Created new point map with colorblind safe colors and new shapes

@@ -45,8 +45,8 @@ wgs1984.proj <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
 # Download US map
 download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_state_20m.zip", dest="./05_analysis_scripts/map_data/us_states.zip", mode="wb") 
 download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_nation_20m.zip", dest="./05_analysis_scripts/map_data/us_nation.zip", mode="wb") 
-unzip("./05_analysis_scripts/map_data/us_states.zip", exdir = "./05_analysis_scripts/map_data/")
-unzip("./05_analysis_scripts/map_data/us_nation.zip", exdir = "./05_analysis_scripts/map_data/")
+unzip("./05_analysis_scripts/map_data/us_states.zip", exdir = "./05_analysis_scripts/map_data")
+unzip("./05_analysis_scripts/map_data/us_nation.zip", exdir = "./05_analysis_scripts/map_data")
 
 # read in the shapefile
 us.states <- readShapePoly("./05_analysis_scripts/map_data/cb_2014_us_state_20m.shp")
@@ -123,7 +123,8 @@ axis(1, at = seq(10, 200, by = 10), las=2)
 
 # Reproject the raster in the us.atlas projection
 us.cond <- projectRaster(us.cond, crs = us.atlas.proj, over = T)
-plot(log(us.cond), main = "Conductivity", col = pal(255))   #terrain.colors(255))  # rev(heat.colors(255))
+plot(log(us.cond), col = pal(255), axes = F, box = F)   #terrain.colors(255))  # rev(heat.colors(255))
+title("Conductivity", line =  -2)
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
@@ -153,7 +154,8 @@ plot(us.alk)
 
 # Reproject the raster in the us.atlas projection
 us.alk <- projectRaster(us.alk, crs = us.atlas.proj, over = T)
-plot(log(us.alk), main = "Alkalinity", col = pal(255))
+plot(log(us.alk), main =, col = pal(255), axes = F, box = F)
+title("Alkalinity", line = -2)
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
 
[…]

### 2017-05-18 05_analysis_scripts/nes_summarystats.R
align plus minus signs, rm n values

@@ -1,21 +1,28 @@
-# This code calculates a mean, standard deviation, and N for each measured
-# variable by region & lake type ##############################################
+# Calculate mean, standard deviation, and N for each measured variable
+# Setup ####
 
 nes_all <- read.csv("nes_data.csv", header=TRUE, na.strings="NA", stringsAsFactors = FALSE)
 
 nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "Northeastern"
-nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "Eastern"
+nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "Southeastern"
 nes_all[nes_all$state %in% c("ARKANSAS", "IOWA", "KANSAS", "LOUISIANA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "OKLAHOMA", "SOUTH DAKOTA", "TEXAS"), "region"] <- "Central"
 nes_all[nes_all$state %in% c("ARIZONA", "CALIFORNIA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON", "WYOMING"), "region"] <- "Western"
 nes_all$region <- as.factor(nes_all$region)
-nes_all$region <- factor(nes_all$region, levels = c("Western", "Central", "Northeastern", "Eastern"))
[…]

### 2017-05-18 05_analysis_scripts/nes_summarystats.R
make a separate n table, closes #11

@@ -188,9 +188,47 @@ reorg_table <- function(dt, include_header = FALSE){
   res
 }
 
+reorg_table_n <- function(dt, include_header = FALSE){
+
+  dt$param <- dt$variable
+  dt <- dt[,-which(names(dt) %in% 
+                     c("variable", "stat", "variable_type", "mean", "sd"))]
+  
+  vars <- as.character(unique(dt$param))
+  
+  dt <- melt(dt, id = c("param", "region"))
+  dt <- dt[order(dt$param, dt$region),]
+  dt <- t(dcast(dt, variable ~ param + region, value.var = "value"))
+  dt <- dt[-1,]
+  
+  res <- matrix(nrow = length(dt) / 4, ncol = 4)
+  
+  # Western
+  res[,1] <- dt[seq(1, length(dt), by = 4)]
+  # Central
+  res[,2] <- dt[seq(2, length(dt), by = 4)]
+  # Northeastern
+  res[,3] <- dt[seq(3, length(dt), by = 4)]
+  # Southeastern
+  res[,4] <- dt[seq(4, length(dt), by = 4)]
+  
+  res <- data.frame(res, stringsAsFactors = FALSE)
+  res$Variable <- vars
+  
+  res <- res[,c(ncol(res), 1:(ncol(res) - 1))]
+  
+  names(res)[2:5] <- c("Western", "Central", "Northeastern", "Southeastern")
+  
+  res
+}
+
 big_reorg <- rbind(reorg_table(morphdf, include_header = TRUE),
                    reorg_table(physchemdf),
                    reorg_table(loaddf))
+
+big_reorg_n <- rbind(reorg_table_n(morphdf),
+                   reorg_table_n(physchemdf),
+                   reorg_table_n(loaddf))
 
 morph_wide    <- dcast(morphdf   , variable ~ region, value.var = c("mean", "sd"))
 physchem_wide <- dcast(physchemdf, variable ~ region, value.var = "stat")  
@@ -249,4 +287,10 @@ sink_text(
                           "Northeastern" = 3,
                           "Southeastern" = 3))), 
   "07_tables/big_table_no_n.tex")
+
+sink_text(
+  (knitr::kable(big_reorg_n, 
+                format = "latex", booktabs = T, row.names = FALSE) %>% 
+     kable_styling(latex_options = "striped")), 
+  "07_tables/big_table_n.tex")
 

### 2017-07-29 README.Rmd
typo in metadata for mean depth

@@
 |surface_area   | the area of the water surface|sq km|
-|mean_depth    | the volume of the water body divided by the surface area in square meters| cubic meters|
+|mean_depth    | the volume of the water body divided by the surface area in square meters| meters|
 |total_inflow      | the mean of the inflows of all tributaries and the immediate drainage | cubic meters per second|

### 2017-07-29 05_analysis_scripts/nes_summarystats.R
add units to table 2

@@ -1,14 +1,30 @@
 # Calculate mean, standard deviation, and N for each measured variable
 # Setup ####
 
-nes_all <- read.csv("nes_data.csv", header=TRUE, na.strings="NA", stringsAsFactors = FALSE)
+nes_all <- read.csv("nes_data.csv", header=TRUE, na.strings="NA", 
+                    stringsAsFactors = FALSE)
 
-nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", "NEW HAMPSHIRE", "NEW YORK", "MAINE", "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", "MICHIGAN"), "region"] <- "Northeastern"
-nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA", "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", "OHIO", "PENNSYLVANIA", "SOUTH CAROLINA", "TENNESSEE", "VIRGINIA", "WEST VIRGINIA"), "region"] <- "Southeastern"
-nes_all[nes_all$state %in% c("ARKANSAS", "IOWA", "KANSAS", "LOUISIANA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "OKLAHOMA", "SOUTH DAKOTA", "TEXAS"), "region"] <- "Central"
-nes_all[nes_all$state %in% c("ARIZONA", "CALIFORNIA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON", "WYOMING"), "region"] <- "Western"
+nes_all[nes_all$state %in% c("VERMONT", "CONNECTICUT", "RHODE ISLAND", 
+                             "NEW HAMPSHIRE", "NEW YORK", "MAINE", 
+                             "MASSACHUSETTS", "WISCONSIN", "MINNESOTA", 
+                             "MICHIGAN"), "region"] <- "Northeastern"
+nes_all[nes_all$state %in% c("ALABAMA", "DELAWARE", "FLORIDA", "GEORGIA",
+                             "ILLINOIS", "INDIANA", "KENTUCKY", "MARYLAND", 
+                             "MISSISSIPPI", "NEW JERSEY", "NORTH CAROLINA", 
[…]

### 2017-07-29 README.Rmd
use updated fig in readme

@@
 library(magick)
-points <- image_read("06_images/points.jpeg")
-image_write(image_trim(points), "06_images/points_trim.jpeg")
+points <- image_read("06_images/20170526/points_newnames_20170530.jpeg")
+image_write(image_trim(points), 
+            "06_images/20170526/points_newnames_20170530_trim.jpeg")
 ```
 
-![](06_images/points_trim.jpeg)
+![](06_images/20170526/points_newnames_20170530_trim.jpeg)
 

### 2017-11-06 08_manuscript/template.tex
add copernicus latex template

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[journal abbreviation, manuscript]{copernicus}

%% Journal abbreviations (please use the same for discussion papers and final revised papers)

% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

\begin{document}

\title{TEXT}

% \Author[affil]{given_name}{surname}

\Author[]{}{}
\Author[]{}{}
\Author[]{}{}

\affil[]{ADDRESS}
\affil[]{ADDRESS}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

\runningtitle{TEXT}

\runningauthor{TEXT}

\correspondence{NAME (EMAIL)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
TEXT
\end{abstract}

\copyrightstatement{TEXT}

\introduction  %% \introduction[modified heading if necessary]
TEXT

\section{HEADING}
TEXT

\subsection{HEADING}
TEXT

\subsubsection{HEADING}
TEXT

\conclusions  %% \conclusions[modified heading if necessary]
TEXT

%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.

\codeavailability{TEXT} %% use this section when having only software code available

\dataavailability{TEXT} %% use this section when having only data sets available

\codedataavailability{TEXT} %% use this section when having data sets and software code available

\appendix
\section{}    %% Appendix A

\subsection{}     %% Appendix A1, A2, etc.

\noappendix       %% use this to mark the end of the appendix section

%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:

%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
%% They will be correctly named automatically.

%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:

\appendixfigures  %% needs to be added in front of appendix figures

\appendixtables   %% needs to be added in front of appendix tables

%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.

\authorcontribution{TEXT} %% optional section

\competinginterests{TEXT} %% this section is mandatory even if you declare that no competing interests are present

\disclaimer{TEXT} %% optional section

\begin{acknowledgements}
TEXT
\end{acknowledgements}

%% REFERENCES

%% The reference list is compiled as follows:

\begin{thebibliography}{}

\bibitem[AUTHOR(YEAR)]{LABEL}
REFERENCE 1

\bibitem[AUTHOR(YEAR)]{LABEL}
REFERENCE 2

\end{thebibliography}

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
%% \bibliographystyle{copernicus}
%% \bibliography{example.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

%% FIGURES

%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
%% between bibliography and first table and/or figure as well as between each table and/or figure.

%% ONE-COLUMN FIGURES

%%f
%\begin{figure}[t]
%\includegraphics[width=8.3cm]{FILE NAME}
%\caption{TEXT}
%\end{figure}
%
%%% TWO-COLUMN FIGURES
%
%%f
%\begin{figure*}[t]
%\includegraphics[width=12cm]{FILE NAME}
%\caption{TEXT}
%\end{figure*}
%
%
%%% TABLES
%%%
%%% The different columns must be seperated with a & command and should
%%% end with \\ to identify the column brake.
%
%%% ONE-COLUMN TABLE
%
%%t
%\begin{table}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table}
%
%%% TWO-COLUMN TABLE
%
%%t
%\begin{table*}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table*}
%
%
%%% MATHEMATICAL EXPRESSIONS
%
%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
%%%
%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
%%% Vectors are identified in bold italic font using \vec{x}
%%% Matrices are identified in bold roman font
%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
%%% The character * should not be applied as mutliplication sign
%
%
%%% EQUATIONS
%
%%% Single-row equation
%
%\begin{equation}
%
%\end{equation}
%
%%% Multiline equation
%
%\begin{align}
%& 3 + 5 = 8\\
%& 3 + 5 = 8\\
%& 3 + 5 = 8
%\end{align}
%
%
%%% MATRICES
%
%\begin{matrix}
%x & y & z\\
%x & y & z\\
%x & y & z\\
%\end{matrix}
%
%
%%% ALGORITHM
%
%\begin{algorithm}
%\caption{…}
%\label{a1}
%\begin{algorithmic}
%…
%\end{algorithmic}
%\end{algorithm}
%
%
%%% CHEMICAL FORMULAS AND REACTIONS
%
%%% For formulas embedded in the text, please use \chem{}
%
%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
%
%\begin{reaction}
%%% \rightarrow should be used for normal (one-way) chemical reactions
%%% \rightleftharpoons should be used for equilibria
%%% \leftrightarrow should be used for resonance structures
%\end{reaction}
%
%
%%% PHYSICAL UNITS
%%%
%%% Please use \unit{} and apply the exponential notation

\end{document}

### 2017-11-06 08_manuscript/template.tex
fill-in template text

@@
 
-\title{TEXT}
+\title{The National Eutrophication Survey: lake characteristics and historical nutrient concentrations}
 
@@
 
-\Author[]{}{}
-\Author[]{}{}
-\Author[]{}{}
+\Author[1]{Joseph}{Stachelek}
+\Author[2]{Chanse}{Ford}
+\Author[3]{Dustin}{Kincaid}
+\Author[1]{Katelyn}{King}
+\Author[4]{Heather}{Miller}
+\Author[5]{Nagelkirk}{Ryan}
 
-\affil[]{ADDRESS}
-\affil[]{ADDRESS}
+\affil[1]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, MI, USA}
+\affil[2]{Department of Earth and Environmental Sciences, Michigan State University, East Lansing, MI, USA}
+\affil[3]{Department of Integrative Biology, Michigan State University, East Lansing, MI, USA}
+\affil[4]{Department of Microbiology and Molecular Genetics, Michigan State University, East Lansing, MI, USA}
+\affil[5]{Department of Geography, Environment, and Spatial Sciences, Michigan State University, East Lansing, MI, USA}
 
@@
 
-\runningtitle{TEXT}
+\runningtitle{National Eutrophication Survey}
 
-\runningauthor{TEXT}
+\runningauthor{Stachelek et al.}
 
-\correspondence{NAME (EMAIL)}
+\correspondence{Joseph Stachelek (stachel2@msu.edu)}
 
@@
 \begin{abstract}
-TEXT
+Historical ecological surveys serve as a baseline and provide context for contemporary research, yet many of these records are not preserved in a way that ensures their long-term usability.
+The National Eutrophication Survey database is currently only available as scans of the original reports (PDF files) with no embedded character information.
+This limits its searchability, machine readability, and the ability of current and future scientists to systematically evaluate its contents.
+These data were collected by the United States Environmental Protection Agency between 1972 and 1975 as part of an effort to investigate eutrophication in freshwater lakes and reservoirs.
+Although several studies have manually transcribed small portions of the database in support of specific studies, there have been no systematic attempts to transcribe and preserve the database in its entirety.
+Here we use a combination of automated optical character recognition and manual quality assurance procedures to make these data available for analysis.
+The performance of the optical character recognition protocol was found to be linked to variation in the quality (clarity) of the original documents.
+For each of the four archival scanned reports, our quality assurance protocol found an error rate between 5.9 and 17\%.
+The goal of our approach was to strike a balance between efficiency and data quality by combining hand-entry of data with digital transcription technologies.
+The finished database contains information on the physical characteristics, hydrology, and water quality of about 800 lakes in the contiguous United States (doi:10.5063/F10G3H3Z).
+Ultimately, this database could be combined with more recent studies to generate metadata analyses of water quality trends and spatial variation across the continental United States.
 \end{abstract}
 
-\copyrightstatement{TEXT}
+% \copyrightstatement{TEXT}
 
 \introduction  %% \introduction[modified heading if necessary]
-TEXT
+Effective management of inland freshwater lakes requires an understanding of the factors that affect water quality and how these factors change over time.
+One of these factors, termed eutrophication, occurs when excess nutrient inputs from human activities fuels increases in algal growth which can cause hypoxia and decreases in water clarity.
+Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture (Smith et al., 1999, Smith et al., 2014).
+As human populations and their impacts continue to grow, eutrophication is expected to become more widespread (Bennett et al., 2001, Taranu and Gregory-Eaves, 2008).
+Historical datasets are needed in order to track, understand, and manage eutrophication in lakes and reservoirs because they serve as an important baseline for modern studies.
 
-\section{HEADING}
-TEXT
+The U.S.
+Environmental Protection Agency (EPA) designed and implemented the National Eutrophication Survey (NES) in order to investigate the extent of eutrophication in freshwater lakes and reservoirs across the contiguous United States (US).
+Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
+Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge (USEPA 2009).
+Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure 1).
+In the remainder of the present paper we refer to the former two regions as simply the northeastern and southeastern regions.
 
-\subsection{HEADING}
-TEXT
+To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.
+For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance (Soballe and Kimmel, 1987).
+Also, it was used to predict eutrophication incidence in a Bayesian framework (Lamon and Stow 2004).
+Smaller portions of the data were used to explore drivers of nutrient loading (Stomp et al., 2011, Brett and Benjamin, 2007).
+Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is Stomp et al., (2011), and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
 
-\subsubsection{HEADING}
-TEXT
+The present study is the first to leverage digital transcription technologies to unlock the full NES dataset.
+In this paper, we describe the digital transcription of the full NES dataset with the goal of making the dataset openly accessible to the research community.
+We introduce and publish the data in an open format that requires no proprietary software.
+It can be easily downloaded, used for analysis, and amended.
+The provided summary statistics and figures also allow users to quickly assess the utility of the data..
+Finally, the code and raw data files are provided to facilitate the extraction of fields not represented in our completed dataset (mostly phytoplankton diversity data).
 
-\conclusions  %% \conclusions[modified heading if necessary]
-TEXT
+\section{Methods}
+
+Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analysis.
+Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.
+More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication (USEPA, 1975).
+Due to historical nature of the dataset, the NES sampling design differs from more modern efforts (USEPA, 2009).
+For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
+As such, NES data values represent the median of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
+
+We obtained the NES archival scanned reports from the EPA National Service Center for Environmental Publications (available at: https://www.epa.gov/nscep).
+The data for each NES region is contained in four separate files.
+We extracted the data from each file using automated techniques followed by manual quality assurance and checking.
+To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/).
+Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) (Ooms, 2017, Smith, 2007).
+The output of these initial extraction steps were recorded in a set of “raw data” files where each file contains the raw unprocessed text of each document page.
+The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package (Stachelek, 2017a).
+Finally, all values in the database were manually checked for accuracy against the original scanned reports.
+Inaccurate OCR outputs were hand-corrected in the final database.
+
+We provide the final dataset in an open non-proprietary format (comma-delimited, *.csv).
+We generated metadata descriptions from the contents of the original scanned reports.
+All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts (R Core Team, 2017).
+Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages (Wickham, 2007, Wickham, 2011, Pebesma and Bivand, 2005).
+
+\section{Results}
+
+The final NES dataset contains observations from 775 lakes and the distribution of these lakes was spatially variable.
+Although there were more lakes measured in the northeastern and southeastern United States, the number of locations was close to evenly distributed among the remaining regions (Figure 1, Table 1).
+Specifically, the number of lakes sampled in each region were as follows: northeastern - 200 lakes, southeastern - 245 lakes, central - 177 lakes, and western, 153 lakes.
+
+Although the overall number of lakes was similar among regions, they differed in the proportion of lakes classified as impoundments rather than natural lakes.
+For example, slightly more than half of the lakes studied (462 of 775) were classified as impoundments yet the northeastern region had only 54 impoundments and the southeastern region had 168 impoundments.
+Conversely, the number of natural lakes sampled in the northeastern region (146 lakes) was more than double that of any other region (77, 48, and 42, for southeastern, western, and central United States, respectively).
+
+We observed substantial spatial variation in many of the individual lake characteristics.
+For example, lakes in the eastern sub-regions were generally smaller and shallower than lakes in the western sub-region (Table 2).
+In addition, lakes in the western sub-region generally had higher alkalinity and higher water clarity (Figure 2, 3).
+Lakes with particularly low alkalinity were found in coastal areas, whereas lakes with particularly high alkalinity were found in Nevada, western Washington, and parts of North Dakota.
+Comparisons among regions was easy for some well-sampled lake chemistry parameters such as total phosphorus but more difficult for undersampled lake chemistry parameters.
+A particularly extreme example of this difficulty was total nitrogen measurements in the eastern region, as this parameter was only measured for a single lake (Table 1).
+
+The ability to examine these spatial trends was made possible by our optical character recognition procedure which had 6 - 17\% accuracy depending on region and archival report scan quality.
+In total, we carried out approximately 5,000 hand-corrections to the automated data product as part of our manual quality control review.
+A total of approximately 650 lakes had values for at least 80\% of the total number of variables shown in Table 1.
+On an individual lake basis, the most common “missing” data was nutrient loading estimates for individual point and nonpoint-source components.
+In many cases, this data may not actually be missing but it may have not been a component of the budget for that particular lake.
+For example, not all lakes have industrial land use so no data is expected in these cases.
+
+\section{Discussion}
+
+We have demonstrated an approach for rescuing historical data from scanned documents.
+In particular, our approach involved a two-step process of automated data scraping followed by hand-curation and quality assurance.
+Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records (Drinkwater et al., 2014).
+Unfortunately, optical character recognition technology does not have absolute accuracy.
+In our case, transcription was hampered by poor print and scan quality of the source paper documents.
+We discovered through our manual validation procedure that the OCR computations produced inaccurate values in approximately 6 - 17\% of the cells in the complete dataset (n = 4836).
+We expect that accuracy could be improved by experimenting with varying the window size of the local adaptive thresholding algorithm relative to the document font size.
+Our ability to experiment with thresholding window size was limited due to the computationally expensive nature of these extractions.
+
+The end result of our approach was data from every lake and nearly every variable in the NES survey dataset.
+The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by Stomp et al.
+(2011).
+The results of the present study could be used to explore anthropogenic and environmental drivers of lake eutrophication as well as to verify previously documented trends.
+One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes (USEPA, 2009).
+This reanalysis considered population level trends in the NES lakes but did not consider trends in individual lakes or potential environmental drivers contributing to observed trends.
+On a population basis, the NLA reanalysis found that less than 30\% of the NES lakes had increased chlorophyll and phosphorus concentrations.
+The results of the present study could be used to verify these claims as well as to compare the NES data with more recent work such as the 2012 National Lakes Assessment.
+Note that sampling techniques may differ from current techniques, so care should be given when making comparisons.
+In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time.
+Such data is critical to a variety of hydrological and water quality modelling efforts (Brett and Benjamin, 2008).
+
+Although our goal was to digitally transcribe the full NES dataset to facilitate studies on historical nutrient loading, it is worth noting the similarities between the present study and other scientific record digitization initiatives.
+Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences (Allan et al., 2011, Freeman et al., 2016).
+To our knowledge, the present study is the first large-scale attempt at digitization of historical limnology records.
+We hope that by making our analysis open and reproducible we will inspire future efforts at recovering important records from the pre-digital era.
 
@@
 
-\codeavailability{TEXT} %% use this section when having only software code available
+% \codeavailability{TEXT} %% use this section when having only software code available
 
-\dataavailability{TEXT} %% use this section when having only data sets available
+% \dataavailability{} %% use this section when having only data sets available
 
-\codedataavailability{TEXT} %% use this section when having data sets and software code available
+\codedataavailability{
+Original scanned reports from the EPA are available from the EPA National Service Center for Environmental Publications (https://www.epa.gov/nscep).
+Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z.
+The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions (Stachelek et al., 2017b).
+Moreover, the R package and R code used to scrape and analyze the data are provided by Stachelek (2017a) so that the methods may be reproduced and openly available for (re)use.
+All figures and summary statistics were generated by R scripts available in the data supplement linked above.
+} %% use this section when having data sets and software code available
 
-\appendix
-\section{}    %% Appendix A
+% \appendix
+% \section{}    %% Appendix A
+% 
+% \subsection{}     %% Appendix A1, A2, etc.
 
-\subsection{}     %% Appendix A1, A2, etc.
-
-\noappendix       %% use this to mark the end of the appendix section
+% \noappendix       %% use this to mark the end of the appendix section
 
@@
 
-\authorcontribution{TEXT} %% optional section
+\authorcontribution{
+All authors contributed to data quality assurance and edited the article text.
+JS conceived the study and implemented the optical character recognition code.
+CF, DK, and RN performed the data analysis and made figures.
+KK, HM, and JS wrote major parts of the manuscript text.
+} %% optional section
 
-\competinginterests{TEXT} %% this section is mandatory even if you declare that no competing interests are present
+\competinginterests{
+The authors declare that they have no conflict of interest.
+} %% this section is mandatory even if you declare that no competing interests are present
 
-\disclaimer{TEXT} %% optional section
+% \disclaimer{TEXT} %% optional section
 
-\begin{acknowledgements}
-TEXT
-\end{acknowledgements}
+% \begin{acknowledgements}
+% TEXT
+% \end{acknowledgements}
 
@@
 
-\begin{thebibliography}{}
-
-\bibitem[AUTHOR(YEAR)]{LABEL}
-REFERENCE 1
-
-\bibitem[AUTHOR(YEAR)]{LABEL}
-REFERENCE 2
-
-\end{thebibliography}
+% \begin{thebibliography}{}
+% 
+% \bibitem[AUTHOR(YEAR)]{LABEL}
+% REFERENCE 1
+% 
+% \bibitem[AUTHOR(YEAR)]{LABEL}
+% REFERENCE 2
+% 
+% \end{thebibliography}
 
@@
 %%
-%% \bibliographystyle{copernicus}
-%% \bibliography{example.bib}
+\bibliographystyle{copernicus}
+\bibliography{bib.bib}
 %%
@@
 %\begin{algorithm}
-%\caption{…}
+%\caption{?}
 %\label{a1}
 %\begin{algorithmic}
-%…
+%?
 %\end{algorithmic}

### 2017-11-06 05_analysis_scripts/maps.R
only need to download once

@@ -8,7 +8,7 @@
 
 # Load Libraries
 library(downloader)
-library(rgdal)
+library(rgdal) # 
 library(sp)
 library(maptools)
 library(ggplot2)
@@ -17,8 +17,6 @@ library(raster)
 library(tmap)
 library(RColorBrewer)
 library(png)
-
-
 
 # Read in the file
 lakes.dat <- read.csv("./nes_data.csv")
@@ -36,9 +34,10 @@ for(i in 1:nrow(lakes.dat)){
   } 
 }
 
-# Map the US
+# Map the US ####
 # CRS for the US National Atlas Map
-us.atlas.proj <- CRS("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs")
+us.atlas.proj <- CRS(paste0("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0", 
+                            " +a=6370997 +b=6370997 +units=m +no_defs"))
 # CRS for the US in WGS 1984 (world geographic 1984)
 wgs1984.proj <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
 

### 2017-11-06 05_analysis_scripts/maps.R
linting

@@ -1,21 +1,13 @@
-# DESCRIPTION
-# This code reads in the Eutrification data table, downloads US base maps and projects
-# the data points and data values, using inverse distance weighting (IDW) for the latter. 
 
-# Written by Ryan Nagelkirk
-# Date: 20170413
-
-
-# Load Libraries
-library(downloader)
+library(downloader) #
 library(rgdal) # 
 library(sp)
-library(maptools)
+library(maptools) # 
 library(ggplot2)
-library(gstat)
-library(raster)
+library(gstat) #
+library(raster) #
 library(tmap)
-library(RColorBrewer)
+library(RColorBrewer) # 
 library(png)
 
 # Read in the file
@@ -43,68 +35,68 @@ wgs1984.proj <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
 
 
 # Download US map
-download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_state_20m.zip", dest="./05_analysis_scripts/map_data/us_states.zip", mode="wb") 
-download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_nation_20m.zip", dest="./05_analysis_scripts/map_data/us_nation.zip", mode="wb") 
-unzip("./05_analysis_scripts/map_data/us_states.zip", exdir = "./05_analysis_scripts/map_data")
-unzip("./05_analysis_scripts/map_data/us_nation.zip", exdir = "./05_analysis_scripts/map_data")
+# download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_state_20m.zip", dest="./05_analysis_scripts/map_data/us_states.zip", mode="wb") 
+# download("http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_nation_20m.zip", dest="./05_analysis_scripts/map_data/us_nation.zip", mode="wb") 
+# unzip("./05_analysis_scripts/map_data/us_states.zip", exdir = "./05_analysis_scripts/map_data")
+# unzip("./05_analysis_scripts/map_data/us_nation.zip", exdir = "./05_analysis_scripts/map_data")
 
 # read in the shapefile
-us.states <- readShapePoly("./05_analysis_scripts/map_data/cb_2014_us_state_20m.shp")
-us.nation <- readShapePoly("./05_analysis_scripts/map_data/cb_2014_us_nation_20m.shp")
[…]

### 2017-11-06 08_manuscript/template.tex
insert inline figures

@@
 
+\begin{figure*}[t]
+  \includegraphics[width=12cm]{points_regions-crop.pdf}
+  \caption{Survey locations colored by sampling year (1972 northeastern-red, 1973 southeastern-green, 1974 central-blue, 1975 western-grey).}
+\end{figure*}
+
 To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.
@@
 
+% insert table 1
+
 Although the overall number of lakes was similar among regions, they differed in the proportion of lakes classified as impoundments rather than natural lakes.
@@
 A particularly extreme example of this difficulty was total nitrogen measurements in the eastern region, as this parameter was only measured for a single lake (Table 1).
+
+% insert table 2
+
+\begin{figure*}[t]
+  \includegraphics[width=12cm]{alkalinity-crop.pdf}
+  \caption{Map of log-scaled alkalinity (mg/L) interpolated using inverse distance weighting.}
+\end{figure*}
+
+\begin{figure*}[t]
+  \includegraphics[width=12cm]{secchi-crop.pdf}
+  \caption{Map of secchi depth (m) interpolated using inverse distance weighting.}
+\end{figure*}
 

### 2017-11-06 05_analysis_scripts/maps.R
pdf generating code, linting

@@ -10,6 +10,14 @@ library(tmap)
 library(RColorBrewer) # 
 library(png)
 
+# https://robjhyndman.com/hyndsight/crop-r-figures/
+savepdf <- function(file, width=16, height=10){
+  fname <- paste("06_images/",file,".pdf",sep="")
+  pdf(fname, width = width, height = height,
+      pointsize = 10)
+  par(mgp=c(2.2, 0.45, 0), tcl = -0.4, mar = c(0, 0, 1.1, 1.1))
+}
+
 # Read in the file
 lakes.dat <- read.csv("./nes_data.csv")
 
@@ -133,12 +141,10 @@ plot(us.cond, add = T)
 # labels=seq(r.range[1], r.range[2], 25), cex.axis=0.6),
 
 ############  ALKALINITY ######################################################
-
-# Make a new df with no NAs for conductivity
 lakes.alk <- lakes.dat[complete.cases(lakes.dat$alkalinity),]
 
-# Apply the IDW
-idw <- idw(formula = alkalinity ~ 1, locations = lakes.alk, newdata = grd)  
+idw     <- idw(formula = alkalinity ~ 1, locations = lakes.alk, 
+               newdata = grd)  
 idw.ras <- raster(idw)
 
 # Rasterize and crop the IDW (might not be needed)
@@ -147,47 +153,46 @@ proj4string(idw.ras) <- wgs1984.proj
 
 # Crop the data to the US outline
 us.alk <- mask(idw.ras, us.unproj.nation)
-plot(us.alk)
+# plot(us.alk)
 
-# Reproject the raster in the us.atlas projection
+savepdf("alkalinity", 5, 4.7)
 us.alk <- projectRaster(us.alk, crs = us.atlas.proj, over = T)
 plot(log(us.alk), col = pal(255), axes = F, box = F, legend = F)
 plot(log(us.alk), legend.only = T, col = pal(255), 
      smallplot = c(.755, .78, .4, .65), 
      legend.args = list(text =  "log(alkalinity)", side = 4, font = 2, 
                         line = 2, cex = .8))
-title("Alkalinity", line = -2)
+# title("Alkalinity", line = -2)
 plot(us.48, add = T, border = "gray")
 plot(us.nation, add = TRUE)
-
+dev.off()
+system("pdfcrop 06_images/alkalinity.pdf")
+unlink("06_images/alkalinity.pdf")
 
[…]

### 2017-11-07 08_manuscript/template.tex
add tables

@@
 \Author[4]{Heather}{Miller}
-\Author[5]{Nagelkirk}{Ryan}
+\Author[5]{Ryan}{Nagelkirk}
 
@@
 
-% insert table 1
+\begin{table*}
+\caption{Number of measurements (n) for each variable in each NES region.}
+\begin{tabular}{lllll}
+\tophline
+Variable & Western & Central & Northeastern & Southeastern\\
+\middlehline
+Drainage area & 122 & 138 & 171 & 232\\
+\hline
+Surface area & 152 & 177 & 200 & 245\\
+
+Mean depth & 149 & 174 & 174 & 242\\
+
+Total inflow & 124 & 138 & 170 & 232\\
+
+Retention time & 124 & 140 & 158 & 230\\
+
+Alkalinity & 153 & 177 & 200 & 245\\
+
+Conductivity & 153 & 176 & 200 & 245\\
+
+Secchi depth & 153 & 177 & 200 & 245\\
+
+Total P & 153 & 177 & 200 & 245\\
+
+Total inorg. P & 153 & 177 & 200 & 245\\
+
+Total inorg. N & 153 & 177 & 200 & 245\\
+
+Total N & 152 & 176 & 1 & 245\\
+
+P pt. source mun. & 52 & 83 & 139 & 189\\
+
+P pt. source ind. & 7 & 1 & 10 & 24\\
+
+P pt. source sep. & 65 & 88 & 111 & 175\\
+
+P nonpt. source & 122 & 133 & 167 & 231\\
+
+P total inputs & 122 & 133 & 167 & 231\\
+
+N pt. source mun. & 52 & 84 & 139 & 189\\
+
+N pt. source ind. & 7 & 1 & 8 & 22\\
+
+N pt. source sep. & 77 & 90 & 111 & 184\\
+
+N nonpt. source & 122 & 129 & 167 & 231\\
+
+N total inputs & 122 & 129 & 167 & 231\\
+
+P total exports & 119 & 132 & 167 & 227\\
+
+P retention & 99 & 115 & 144 & 201\\
+
+P load per area & 122 & 133 & 167 & 231\\
+
+N total exports & 119 & 133 & 166 & 227\\
+
+N retention & 88 & 111 & 122 & 170\\
+
+N load per area & 122 & 135 & 167 & 231\\
+\bottomhline
+\end{tabular}
+\end{table*}
+
+%\begin{table*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table*}
 
@@
 
-% insert table 2
+\begin{table*}[t]
+\caption{Mean and standard deviation (sd) for each variable in each NES region.}
+\begin{tabular}{lllllllllllll}
+\tophline
+\multicolumn{1}{l|}{Region} & \multicolumn{3}{|c|}{Western} & \multicolumn{3}{|c|}{Central} & \multicolumn{3}{|c|}{Northeastern} & \multicolumn{3}{|c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
+Variable & Mean &   & sd & Mean &   & sd & Mean &   & sd & Mean &   & sd\\
+\middlehline
+Drainage area ($km^{2}$) & 2.5e+04 & ± & 7.8e+04 & 2.1e+04 & ± & 7.5e+04 & 3.2e+03 & ± & 1.4e+04 & 5.3e+03 & ± & 1.4e+04\\
+
+Surface area ($km^{2}$) & 44.57 & ± & 99.83 & 54.38 & ± & 1.4e+02 & 27.25 & ± & 99.01 & 42.7 & ± & 1.4e+02\\
+
+Mean depth ($m$) & 16.71 & ± & 27.08 & 5.97 & ± & 4.49 & 7 & ± & 9.37 & 6.4 & ± & 6.07\\
+
+Total inflow ($m^{3} \cdot s^{-1}$) & 52.1 & ± & 1.1e+02 & 31.82 & ± & 71.77 & 23.1 & ± & 65.26 & 82.6 & ± & 2.3e+02\\
+
+Retention time ($years$) & 7.27 & ± & 43.32 & 2.78 & ± & 6.98 & 2.01 & ± & 4.77 & 0.59 & ± & 1.12\\
+
+Alkalinity ($mg \cdot l^{-1}$) & 1.7e+02 & ± & 3.7e+02 & 1.5e+02 & ± & 91.51 & 1.2e+02 & ± & 1.6e+02 & 72.18 & ± & 66.25\\
+
+Conductivity ($uohm$) & 4.9e+02 & ± & 1.0e+03 & 6.4e+02 & ± & 7.6e+02 & 3.3e+02 & ± & 4.0e+02 & 2.5e+02 & ± & 2.2e+02\\
+
+Secchi depth ($m$) & 2.86 & ± & 2.64 & 1.2 & ± & 0.91 & 1.81 & ± & 1.71 & 1.22 & ± & 0.82\\
+
+Total P ($mg \cdot l^{-1}$) & 0.07 & ± & 0.13 & 0.11 & ± & 0.16 & 0.16 & ± & 0.35 & 0.12 & ± & 0.27\\
+
+Total inorg. P ($mg \cdot l^{-1}$) & 0.04 & ± & 0.11 & 0.04 & ± & 0.07 & 0.11 & ± & 0.3 & 0.05 & ± & 0.15\\
+
+Total inorg. N ($mg \cdot l^{-1}$) & 0.14 & ± & 0.23 & 0.33 & ± & 0.58 & 0.47 & ± & 0.66 & 0.72 & ± & 0.91\\
+
+Total N ($mg \cdot l^{-1}$) & 0.62 & ± & 0.65 & 1.22 & ± & 1.11 & 0.12 & ± & NA & 1.56 & ± & 1.25\\
+
+P pt. source mun. ($kg \cdot yr^{-1}$) & 2.5e+04 & ± & 8.7e+04 & 2.3e+04 & ± & 5.6e+04 & 3.5e+04 & ± & 1.5e+05 & 4.5e+04 & ± & 1.1e+05\\
+
+P pt. source ind. ($kg \cdot yr^{-1}$) & 2.5e+04 & ± & 4.0e+04 & 1.3e+04 & ± & NA & 2.7e+04 & ± & 4.9e+04 & 1.7e+04 & ± & 4.5e+04\\
+
+P pt. source sep. ($kg \cdot yr^{-1}$) & 56.62 & ± & 1.4e+02 & 60.62 & ± & 93.67 & 1.6e+02 & ± & 3.4e+02 & 98.55 & ± & 2.3e+02\\
+
+P nonpt. source ($kg \cdot yr^{-1}$) & 1.4e+05 & ± & 4.2e+05 & 1.8e+05 & ± & 6.8e+05 & 5.6e+04 & ± & 2.1e+05 & 1.9e+05 & ± & 5.5e+05\\
+
+P total inputs ($kg \cdot yr^{-1}$) & 1.5e+05 & ± & 4.7e+05 & 2.0e+05 & ± & 7.0e+05 & 8.7e+04 & ± & 3.4e+05 & 2.3e+05 & ± & 5.8e+05\\
+
+N pt. source mun. ($kg \cdot yr^{-1}$) & 7.8e+04 & ± & 2.5e+05 & 7.3e+04 & ± & 1.7e+05 & 1.4e+05 & ± & 5.4e+05 & 1.4e+05 & ± & 3.8e+05\\
+
+N pt. source ind. ($kg \cdot yr^{-1}$) & 2.3e+07 & ± & 6.1e+07 & 4.0e+03 & ± & NA & 1.6e+05 & ± & 4.2e+05 & 1.7e+05 & ± & 5.6e+05\\
+
+N pt. source sep. ($kg \cdot yr^{-1}$) & 5.7e+06 & ± & 5.0e+07 & 2.2e+03 & ± & 3.5e+03 & 4.3e+03 & ± & 5.5e+03 & 3.3e+03 & ± & 6.7e+03\\
+
+N nonpt. source ($kg \cdot yr^{-1}$) & 1.8e+06 & ± & 4.9e+06 & 1.8e+06 & ± & 4.4e+06 & 1.2e+06 & ± & 4.1e+06 & 3.1e+06 & ± & 8.9e+06\\
+
+N total inputs ($kg \cdot yr^{-1}$) & 6.8e+06 & ± & 5.7e+07 & 1.8e+06 & ± & 4.3e+06 & 1.3e+06 & ± & 4.6e+06 & 3.2e+06 & ± & 9.0e+06\\
+
+P total exports ($kg \cdot yr^{-1}$) & 6.2e+04 & ± & 1.7e+05 & 7.4e+04 & ± & 1.9e+05 & 7.3e+04 & ± & 3.1e+05 & 1.9e+05 & ± & 6.3e+05\\
+
+P retention (\%) & 47.77 & ± & 28.5 & 57.55 & ± & 26.01 & 36.93 & ± & 25.2 & 42.7 & ± & 23.34\\
+
+P load per area($g / m^{2} / day$) & 5.61 & ± & 21.36 & 3.3 & ± & 9.2 & 28.46 & ± & 97.49 & 9.43 & ± & 17.06\\
+
+N total exports ($kg \cdot yr^{-1}$) & 1.6e+06 & ± & 4.0e+06 & 1.2e+06 & ± & 2.8e+06 & 1.2e+06 & ± & 4.9e+06 & 3.0e+06 & ± & 8.3e+06\\
+
+N retention (\%) & 39.33 & ± & 27.13 & 43.41 & ± & 23.97 & 28.41 & ± & 23.62 & 26.28 & ± & 18.85\\
+
+N load per area ($g / m^{2} / day$) & 1.8e+02 & ± & 1.1e+03 & 42.67 & ± & 1.1e+02 & 2.8e+02 & ± & 9.1e+02 & 1.3e+02 & ± & 2.4e+02\\
+\bottomhline
+\end{tabular}
+\end{table*}
 

### 2017-11-07 05_analysis_scripts/nes_summarystats.R

@@ -309,20 +309,14 @@ units <- c(NA, " ($km^{2}$)", " ($km^{2}$)", " ($m$)",
            " ($g / m^{2} / day$)")
 big_reorg[,"Variable"] <- paste0(big_reorg$Variable, units)
 
-sink_text(
-    (knitr::kable(big_reorg[-1,], 
-        format = "latex", booktabs = T, row.names = FALSE, escape = FALSE) %>% 
-       kable_styling(latex_options = "striped") %>% 
-       add_header_above(c("Region", 
+knitr::kable(big_reorg[-1,], 
+        format = "latex", booktabs = FALSE, row.names = FALSE, escape = FALSE) %>%     add_header_above(c("Region", 
                           "Western" = 3,
                           "Central" = 3,
                           "Northeastern" = 3,
                           "Southeastern" = 3))), 
   "07_tables/big_table_no_n.tex")
 
-sink_text(
-  (knitr::kable(big_reorg_n, 
-                format = "latex", booktabs = T, row.names = FALSE) %>% 
-     kable_styling(latex_options = "striped")), 
-  "07_tables/big_table_n.tex")
+knitr::kable(big_reorg_n, 
+                format = "latex", booktabs = FALSE, row.names = FALSE)
 

### 2017-11-07 08_manuscript/template.tex
add bibtex

@@
 %% 2-column papers and discussion papers
-\documentclass[journal abbreviation, manuscript]{copernicus}
+\documentclass[journal abbreviations, manuscript]{copernicus}
 
@@
 One of these factors, termed eutrophication, occurs when excess nutrient inputs from human activities fuels increases in algal growth which can cause hypoxia and decreases in water clarity.
-Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture (Smith et al., 1999, Smith et al., 2014).
-As human populations and their impacts continue to grow, eutrophication is expected to become more widespread (Bennett et al., 2001, Taranu and Gregory-Eaves, 2008).
+Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture \citep{smith1999eutrophication, smith2014comment}.
+As human populations and their impacts continue to grow, eutrophication is expected to become more widespread \citep{bennett2001human, taranu2008quantifying}.
 Historical datasets are needed in order to track, understand, and manage eutrophication in lakes and reservoirs because they serve as an important baseline for modern studies.
@@
 Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
-Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge (USEPA 2009).
+Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}.
 Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure 1).
@@
 To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.
-For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance (Soballe and Kimmel, 1987).
+For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance \citep{soballe1987large}.
 Also, it was used to predict eutrophication incidence in a Bayesian framework (Lamon and Stow 2004).
-Smaller portions of the data were used to explore drivers of nutrient loading (Stomp et al., 2011, Brett and Benjamin, 2007).
-Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is Stomp et al., (2011), and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
+Smaller portions of the data were used to explore drivers of nutrient loading \citep{stomp2011large,Brettreviewreassessmentlake2007}.
+Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is \citet{stomp2011large}, and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
 
@@
 Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.
-More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication (USEPA, 1975).
-Due to historical nature of the dataset, the NES sampling design differs from more modern efforts (USEPA, 2009).
+More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication \citep{nes-methods}.
+Due to historical nature of the dataset, the NES sampling design differs from more modern efforts \citep{nla-methods}.
 For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
@@
 To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/).
-Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) (Ooms, 2017, Smith, 2007).
+Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) \citep{R-tesseract, smith2007overview}.
 The output of these initial extraction steps were recorded in a set of “raw data” files where each file contains the raw unprocessed text of each document page.
-The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package (Stachelek, 2017a).
+The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package \citep{R-nesR}.
 Finally, all values in the database were manually checked for accuracy against the original scanned reports.
@@
 We generated metadata descriptions from the contents of the original scanned reports.
-All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts (R Core Team, 2017).
-Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages (Wickham, 2007, Wickham, 2011, Pebesma and Bivand, 2005).
+All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts \citep{R-core}.
+Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages \citep{R-plyr,R-sp}.
 
@@
 In particular, our approach involved a two-step process of automated data scraping followed by hand-curation and quality assurance.
-Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records (Drinkwater et al., 2014).
+Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records \citep{drinkwater2014use}.
 Unfortunately, optical character recognition technology does not have absolute accuracy.
@@
 The end result of our approach was data from every lake and nearly every variable in the NES survey dataset.
-The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by Stomp et al.
-(2011).
+The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by \citet{stomp2011large}.
 The results of the present study could be used to explore anthropogenic and environmental drivers of lake eutrophication as well as to verify previously documented trends.
-One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes (USEPA, 2009).
+One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes \citep{nla-methods}.
 This reanalysis considered population level trends in the NES lakes but did not consider trends in individual lakes or potential environmental drivers contributing to observed trends.
@@
 In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time.
-Such data is critical to a variety of hydrological and water quality modelling efforts (Brett and Benjamin, 2008).
+Such data is critical to a variety of hydrological and water quality modelling efforts \citep{Brettreviewreassessmentlake2007}.
 
 Although our goal was to digitally transcribe the full NES dataset to facilitate studies on historical nutrient loading, it is worth noting the similarities between the present study and other scientific record digitization initiatives.
-Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences (Allan et al., 2011, Freeman et al., 2016).
+Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences \citep{allan2011international, freeman2017icoads}.
 To our knowledge, the present study is the first large-scale attempt at digitization of historical limnology records.
@@
 Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z.
-The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions (Stachelek et al., 2017b).
-Moreover, the R package and R code used to scrape and analyze the data are provided by Stachelek (2017a) so that the methods may be reproduced and openly available for (re)use.
+The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions \citep{nes}.
+Moreover, the R package and R code used to scrape and analyze the data are provided by \citet{R-nesR} so that the methods may be reproduced and openly available for (re)use.
 All figures and summary statistics were generated by R scripts available in the data supplement linked above.

### 2017-11-07 08_manuscript/reviewer_response.tex
add response to reviewers

\documentclass{article}
    % General document formatting
    \usepackage[margin=0.7in]{geometry}
    \usepackage[parfill]{parskip}
    \usepackage[utf8]{inputenc}

    % Related to math
    \usepackage{amsmath,amssymb,amsfonts,amsthm}

\begin{document}

\section*{Part a}

Reviewer 1

This is an important data set and it is good to see that it is becoming available in electronic format. The authors appear to have used good methods to bring in a large dataset. I suspect that the manual entry error rate would be close to the lower end of their error rate.

There are multiple copies of these reports floating around in various libraries. Would it improve things to scan multiple copies of the same data and check the copies against each other?

This is a good suggestion. However, the effort to do this would be quite large.  Also, since we manually verified the accuracy of every field, this task seems unnecessary.  This may have been a more useful step at the beginning of our optical character recognition quality control step, affording us perhaps two other scan options, and the best of three could have been populated into our final repository.  We would still suspect that a manual quality control step would have been necessary.

I am curious why the authors did not try to bring in the data that Stomp et al. (2011) digitized. They could then compare the two datasets. At a minimum they should try to combine the datasets, at least chlorophyll would be nice.

The R code provided in the article supplement could be extended to accomplish this task. It would require writing a parsing function to extract the values from the raw files produced as a result of the optical character recognition algorithm. However, this section has very complex formatting so we felt that doing so would be redundant. In addition, the Stomp et al. dataset is already archived and we felt that it was not good data management practice to duplicate their data in a second location.

I am not functional in R so could not assess that part of the data product.

The data could use some quality checking. For example, there are points where phosphate exceeds total phosphorus. This is not possible.

We also agree that some data points reported through these four EPA documents did not make scientific sense.  However, our goal was to reproduce the dataset exactly as it appears in the original documents, and not to judge the scientific accuracy of these data.  We expect that a proper analysis would be conducted in cooperation with an original research project. We have added text to the manuscript pointing out this issue.

It also would be good if the all data were all quality checked. If it takes about 1 second per data point, I calculate it would take 3 hours each for the team of authors working in pairs to check the whole thing. That would lead to a cleaner data set as well as making the error rate certain.

We agree that manual quality checking is essential. We were able to calculate (and report) our error rate because we had already done this manual checking.

It would be nice to have retention all in one type of units (not years or days mixed)

We agree. However, if we converted the data in this way it would make it more difficult to check the data against the original documents.

Table 2. Could use the units

Good catch! We have updated the table.

Reviewer 2

Please use the full doi designation: https://doi.org/10.5063/f1kk98r5 ? The full designation allowed this reviewer to avoid a search through DataCite to access the KNB Site.

The ESSD author instructions say to report a DOI in the abbreviation form (10.5194/xyz). We suspect that this could be hyperlinked to the full designation on typesetting. For example, it appears that the partial doi code provided in the abstract of the pdf manuscript has been hyperlinked to the full doi in the html abstract.

Data very well organized and easily accessible. Very good metadata. Spot checks (Montana, Illinois) showed believable locations and values, evidently quality control has worked reasonably well. Good product, potentially very useful as baseline for both chemical and hydrological / geomorphological purposes.

No information about sampling date in the master .csv file? E.g. a reader gets reference to the report number (.pdf 475, published 1978) and to a page number (for Bloomington Lake, Maclean County, Illinois, actually on page 79 rather than 81 as in .csv file), but no reference to sampling dates. Bloomington Lake data shows nutrient and biological samples collected on 5/11/73, 8/9/73 and 10/17/73. For MacDonald Lake (.pdf 477, page 78 rather than 80 as in .csv) Montana, nutrient and biological sampling on 6/1/75 and 7/28/75. In text we read that sampling of geographic regions occurred by year (e.g. 1973 for southeastern including Illinois and 1975 for western) but the user does not see actual dates where available, or would need to extract those dates themselves? But apparently none of the raw files captured these sampling dates from the original .pdf?

The sampling dates you see are exclusive to the “Biological Characteristics” section, which we did not transcribe, because this has already been done in Stomp et al. (2011). In contrast, the data we report are annual means computed from monthly samples. The NES reports provide no further details about specific sampling dates. We have added clarifying text to the manuscript and metadata on this point.

Data from Illinois resides in two separate sections of .pdf 475 (page numbers 80 to 99 and 100 to 110 contiguous) but one needs to search by storet\_code or state name to find all data per each state? This scattering arises from processing sequence?

Yes

Page numbers in .csv file refer to page number of digitized .pdf, not to page numbers used within the individual reports? I did not see reference to this small discrepancy in the metadata.

We have appended additional text to the “pagenum” metadata field. It not reads: “page number of the pdf (not the report page number)”.

Need specific clarification about the page number discrepancies and about whether the digitisation process captured the sampling date

See above.

\end{document}

### 2017-11-07 08_manuscript/post_response.tex
add response to reviewers

--- 2017-11-07 08_manuscript/template.tex
@@
 Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
+Each lake was sampled for one year on a monthly basis.
+Data is provided in the original NES reports as annual averages.
+Except for the phytoplankton distribution subset, which we did not transcribe \citep[see ][]{stomp2011large}, the NES documents do not provide specific sampling dates.
 Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}.
-Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure 1).
+Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure \ref{fig:points_regions}).
 In the remainder of the present paper we refer to the former two regions as simply the northeastern and southeastern regions.
@@
   \includegraphics[width=12cm]{points_regions-crop.pdf}
-  \caption{Survey locations colored by sampling year (1972 northeastern-red, 1973 southeastern-green, 1974 central-blue, 1975 western-grey).}
+  \caption{Survey locations colored by sampling year (1972 northeastern-red, 1973 southeastern-green, 1974 central-blue, 1975 western-grey).}\label{fig:points_regions}
 \end{figure*}
@@
 For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
-As such, NES data values represent the median of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
+As such, NES data values represent the mean of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
 
@@
 The final NES dataset contains observations from 775 lakes and the distribution of these lakes was spatially variable.
-Although there were more lakes measured in the northeastern and southeastern United States, the number of locations was close to evenly distributed among the remaining regions (Figure 1, Table 1).
+Although there were more lakes measured in the northeastern and southeastern United States, the number of locations was close to evenly distributed among the remaining regions (Figure \ref{fig:points_regions}, Table \ref{table:n}).
 Specifically, the number of lakes sampled in each region were as follows: northeastern - 200 lakes, southeastern - 245 lakes, central - 177 lakes, and western, 153 lakes.
@@
 \begin{table*}
-\caption{Number of measurements (n) for each variable in each NES region.}
+\caption{Number of measurements (n) for each variable in each NES region.}\label{table:n}
 \begin{tabular}{lllll}
@@
 Drainage area & 122 & 138 & 171 & 232\\
-\hline
+
 Surface area & 152 & 177 & 200 & 245\\
@@
 We observed substantial spatial variation in many of the individual lake characteristics.
-For example, lakes in the eastern sub-regions were generally smaller and shallower than lakes in the western sub-region (Table 2).
-In addition, lakes in the western sub-region generally had higher alkalinity and higher water clarity (Figure 2, 3).
+For example, lakes in the eastern sub-regions were generally smaller and shallower than lakes in the western sub-region (Table \ref{table:means}).
+In addition, lakes in the western sub-region generally had higher alkalinity and higher water clarity (Figure \ref{fig:alkalinity}, \ref{fig:secchi}).
 Lakes with particularly low alkalinity were found in coastal areas, whereas lakes with particularly high alkalinity were found in Nevada, western Washington, and parts of North Dakota.
 Comparisons among regions was easy for some well-sampled lake chemistry parameters such as total phosphorus but more difficult for undersampled lake chemistry parameters.
-A particularly extreme example of this difficulty was total nitrogen measurements in the eastern region, as this parameter was only measured for a single lake (Table 1).
+A particularly extreme example of this difficulty was total nitrogen measurements in the eastern region, as this parameter was only measured for a single lake (Table \ref{table:n}).
 
 \begin{table*}[t]
-\caption{Mean and standard deviation (sd) for each variable in each NES region.}
+\caption{Mean and standard deviation (sd) for each variable in each NES region.}\label{table:means}
 \begin{tabular}{lllllllllllll}
 \tophline
-\multicolumn{1}{l|}{Region} & \multicolumn{3}{|c|}{Western} & \multicolumn{3}{|c|}{Central} & \multicolumn{3}{|c|}{Northeastern} & \multicolumn{3}{|c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
+\multicolumn{1}{l}{Region} & \multicolumn{3}{c}{Western} & \multicolumn{3}{c}{Central} & \multicolumn{3}{c}{Northeastern} & \multicolumn{3}{c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
 Variable & Mean &   & sd & Mean &   & sd & Mean &   & sd & Mean &   & sd\\
@@
   \includegraphics[width=12cm]{alkalinity-crop.pdf}
-  \caption{Map of log-scaled alkalinity (mg/L) interpolated using inverse distance weighting.}
+  \caption{Map of log-scaled alkalinity (mg/L) interpolated using inverse distance weighting.}\label{fig:alkalinity}
 \end{figure*}
@@
   \includegraphics[width=12cm]{secchi-crop.pdf}
-  \caption{Map of secchi depth (m) interpolated using inverse distance weighting.}
+  \caption{Map of secchi depth (m) interpolated using inverse distance weighting.}\label{fig:secchi}
 \end{figure*}
@@
 In total, we carried out approximately 5,000 hand-corrections to the automated data product as part of our manual quality control review.
-A total of approximately 650 lakes had values for at least 80\% of the total number of variables shown in Table 1.
+A total of approximately 650 lakes had values for at least 80\% of the total number of variables shown in Table \ref{table:n}.
 On an individual lake basis, the most common “missing” data was nutrient loading estimates for individual point and nonpoint-source components.
@@
 Note that sampling techniques may differ from current techniques, so care should be given when making comparisons.
+For example, 5 lakes have phosphate ($PO^4$) values exceeding their corresponding total phosphorus ($TP$) values.
+This is not physically possible because $PO^4$ is a component of $TP$.
 In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time.

### 2017-11-07 08_manuscript/pre_response.tex
add response to reviewers

--- 2017-11-07 08_manuscript/post_response.tex
@@
 Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
-Each lake was sampled for one year on a monthly basis.
-Data is provided in the original NES reports as annual averages.
-Except for the phytoplankton distribution subset, which we did not transcribe \citep[see ][]{stomp2011large}, the NES documents do not provide specific sampling dates.
 Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}.
@@
 For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
-As such, NES data values represent the mean of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
+As such, NES data values represent the median of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
 
@@
 Drainage area & 122 & 138 & 171 & 232\\
-
+\hline
 Surface area & 152 & 177 & 200 & 245\\
@@
 \tophline
-\multicolumn{1}{l}{Region} & \multicolumn{3}{c}{Western} & \multicolumn{3}{c}{Central} & \multicolumn{3}{c}{Northeastern} & \multicolumn{3}{c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
+\multicolumn{1}{l|}{Region} & \multicolumn{3}{|c|}{Western} & \multicolumn{3}{|c|}{Central} & \multicolumn{3}{|c|}{Northeastern} & \multicolumn{3}{|c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
 Variable & Mean &   & sd & Mean &   & sd & Mean &   & sd & Mean &   & sd\\
@@
 Note that sampling techniques may differ from current techniques, so care should be given when making comparisons.
-For example, 5 lakes have phosphate ($PO^4$) values exceeding their corresponding total phosphorus ($TP$) values.
-This is not physically possible because $PO^4$ is a component of $TP$.
 In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time.

### 2017-11-07 08_manuscript/reviewer_response.tex
append reviewer response

@@
 
-\section*{Part a}
+\section*{Reviewer Responses}
 
-Reviewer 1
+\subsection{Reviewer 1}
 
-This is an important data set and it is good to see that it is becoming available in electronic format.
+\begin{itemize}
+
+    \item \emph{This is an important data set and it is good to see that it is becoming available in electronic format.
 The authors appear to have used good methods to bring in a large dataset.
-I suspect that the manual entry error rate would be close to the lower end of their error rate.
+I suspect that the manual entry error rate would be close to the lower end of their error rate.}
 
-There are multiple copies of these reports floating around in various libraries. Would it improve things to scan multiple copies of the same data and check the copies against each other?
+    \item \emph{There are multiple copies of these reports floating around in various libraries.
+Would it improve things to scan multiple copies of the same data and check the copies against each other?}
 
-This is a good suggestion.
+\textbf{This is a good suggestion.
 However, the effort to do this would be quite large.
@@
 This may have been a more useful step at the beginning of our optical character recognition quality control step, affording us perhaps two other scan options, and the best of three could have been populated into our final repository.
-We would still suspect that a manual quality control step would have been necessary.
+We would still suspect that a manual quality control step would have been necessary.}
 
-I am curious why the authors did not try to bring in the data that Stomp et al.
+    \item \emph{I am curious why the authors did not try to bring in the data that Stomp et al.
 (2011) digitized.
 They could then compare the two datasets.
-At a minimum they should try to combine the datasets, at least chlorophyll would be nice.
+At a minimum they should try to combine the datasets, at least chlorophyll would be nice.}
 
-The R code provided in the article supplement could be extended to accomplish this task.
+\textbf{The R code provided in the article supplement could be extended to accomplish this task.
 It would require writing a parsing function to extract the values from the raw files produced as a result of the optical character recognition algorithm.
@@
 In addition, the Stomp et al.
-dataset is already archived and we felt that it was not good data management practice to duplicate their data in a second location.
+dataset is already archived and we felt that it was not good data management practice to duplicate their data in a second location.}
 
-I am not functional in R so could not assess that part of the data product.
+    \item \emph{I am not functional in R so could not assess that part of the data product.}
 
-The data could use some quality checking. For example, there are points where phosphate exceeds total phosphorus. This is not possible.
+    \item \emph{The data could use some quality checking. For example, there are points where phosphate exceeds total phosphorus. This is not possible.}
 
-We also agree that some data points reported through these four EPA documents did not make scientific sense.
+\textbf{We also agree that some data points reported through these four EPA documents did not make scientific sense.
 However, our goal was to reproduce the dataset exactly as it appears in the original documents, and not to judge the scientific accuracy of these data.
 We expect that a proper analysis would be conducted in cooperation with an original research project.
-We have added text to the manuscript pointing out this issue.
+We have added text to the manuscript pointing out this issue.}
 
-It also would be good if the all data were all quality checked.
+    \item \emph{It also would be good if the all data were all quality checked.
 If it takes about 1 second per data point, I calculate it would take 3 hours each for the team of authors working in pairs to check the whole thing.
-That would lead to a cleaner data set as well as making the error rate certain.
+That would lead to a cleaner data set as well as making the error rate certain.}
 
-We agree that manual quality checking is essential. We were able to calculate (and report) our error rate because we had already done this manual checking.
+\textbf{We agree that manual quality checking is essential. We were able to calculate (and report) our error rate because we had already done this manual checking.}
 
-It would be nice to have retention all in one type of units (not years or days mixed)
+    \item \emph{It would be nice to have retention all in one type of units (not years or days mixed)}
 
-We agree. However, if we converted the data in this way it would make it more difficult to check the data against the original documents.
+\textbf{We agree. However, if we converted the data in this way it would make it more difficult to check the data against the original documents.}
 
-Table 2. Could use the units
+    \item \emph{Table 2. Could use the units}
 
-Good catch! We have updated the table.
+\textbf{Good catch! We have updated the table.}
 
-Reviewer 2
+\end{itemize}
 
-Please use the full doi designation: https://doi.org/10.5063/f1kk98r5 ? The full designation allowed this reviewer to avoid a search through DataCite to access the KNB Site.
+\subsection{Reviewer 2}
 
-The ESSD author instructions say to report a DOI in the abbreviation form (10.5194/xyz).
+\begin{itemize}
+
+    \item \emph{Please use the full doi designation: https://doi.org/10.5063/f1kk98r5 ? The full designation allowed this reviewer to avoid a search through DataCite to access the KNB Site.}
+
+\textbf{The ESSD author instructions say to report a DOI in the abbreviation form (10.5194/xyz).
 We suspect that this could be hyperlinked to the full designation on typesetting.
-For example, it appears that the partial doi code provided in the abstract of the pdf manuscript has been hyperlinked to the full doi in the html abstract.
+For example, it appears that the partial doi code provided in the abstract of the pdf manuscript has been hyperlinked to the full doi in the html abstract.}
 
-Data very well organized and easily accessible.
+\item \emph{Data very well organized and easily accessible.
 Very good metadata.
 Spot checks (Montana, Illinois) showed believable locations and values, evidently quality control has worked reasonably well.
-Good product, potentially very useful as baseline for both chemical and hydrological / geomorphological purposes.
+Good product, potentially very useful as baseline for both chemical and hydrological / geomorphological purposes.}
 
-No information about sampling date in the master .csv file?
+\item \emph{No information about sampling date in the master .csv file?
 E.g.
@@
 1973 for southeastern including Illinois and 1975 for western) but the user does not see actual dates where available, or would need to extract those dates themselves?
-But apparently none of the raw files captured these sampling dates from the original .pdf?
+But apparently none of the raw files captured these sampling dates from the original .pdf?}
 
-The sampling dates you see are exclusive to the “Biological Characteristics” section, which we did not transcribe, because this has already been done in Stomp et al.
+\textbf{The sampling dates you see are exclusive to the “Biological Characteristics” section, which we did not transcribe, because this has already been done in Stomp et al.
 (2011).
@@
 The NES reports provide no further details about specific sampling dates.
-We have added clarifying text to the manuscript and metadata on this point.
+We have added clarifying text to the manuscript and metadata on this point.}
 
-Data from Illinois resides in two separate sections of .pdf 475 (page numbers 80 to 99 and 100 to 110 contiguous) but one needs to search by storet\_code or state name to find all data per each state?
-This scattering arises from processing sequence?
+\item \emph{Data from Illinois resides in two separate sections of .pdf 475 (page numbers 80 to 99 and 100 to 110 contiguous) but one needs to search by storet\_code or state name to find all data per each state?
+This scattering arises from processing sequence?}
 
-Yes
+\textbf{Yes}
 
-Page numbers in .csv file refer to page number of digitized .pdf, not to page numbers used within the individual reports? I did not see reference to this small discrepancy in the metadata.
+\item \emph{Page numbers in .csv file refer to page number of digitized .pdf, not to page numbers used within the individual reports?
+I did not see reference to this small discrepancy in the metadata.}
 
-We have appended additional text to the “pagenum” metadata field. It not reads: “page number of the pdf (not the report page number)”.
+\textbf{We have appended additional text to the “pagenum” metadata field. It not reads: “page number of the pdf (not the report page number)”.}
 
-Need specific clarification about the page number discrepancies and about whether the digitisation process captured the sampling date
+\item \emph{Need specific clarification about the page number discrepancies and about whether the digitisation process captured the sampling date}
 
-See above.
+\textbf{See above.}
+
+\end{itemize}
 

### 2017-11-07 08_manuscript/post_response.tex
check text for flow issues and typos

@@
 Historical ecological surveys serve as a baseline and provide context for contemporary research, yet many of these records are not preserved in a way that ensures their long-term usability.
-The National Eutrophication Survey database is currently only available as scans of the original reports (PDF files) with no embedded character information.
+The National Eutrophication Survey (NES) database is currently only available as scans of the original reports (PDF files) with no embedded character information.
 This limits its searchability, machine readability, and the ability of current and future scientists to systematically evaluate its contents.
-These data were collected by the United States Environmental Protection Agency between 1972 and 1975 as part of an effort to investigate eutrophication in freshwater lakes and reservoirs.
+The NES data were collected by the United States Environmental Protection Agency between 1972 and 1975 as part of an effort to investigate eutrophication in freshwater lakes and reservoirs.
 Although several studies have manually transcribed small portions of the database in support of specific studies, there have been no systematic attempts to transcribe and preserve the database in its entirety.
@@
 The finished database contains information on the physical characteristics, hydrology, and water quality of about 800 lakes in the contiguous United States (doi:10.5063/F10G3H3Z).
-Ultimately, this database could be combined with more recent studies to generate metadata analyses of water quality trends and spatial variation across the continental United States.
+Ultimately, this database could be combined with more recent studies to generate meta-analyses of water quality trends and spatial variation across the continental United States.
 \end{abstract}
@@
 Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
-Each lake was sampled for one year on a monthly basis.
-Data is provided in the original NES reports as annual averages.
-Except for the phytoplankton distribution subset, which we did not transcribe \citep[see ][]{stomp2011large}, the NES documents do not provide specific sampling dates.
+Each lake was sampled on a monthly basis for a period one year.
+Except for the phytoplankton distribution subset, which we did not transcribe \citep[see ][]{stomp2011large}, the NES data is provided as annual averages.
 Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}.
@@
 Smaller portions of the data were used to explore drivers of nutrient loading \citep{stomp2011large,Brettreviewreassessmentlake2007}.
-Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is \citet{stomp2011large}, and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
+Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is \citet{stomp2011large}, but their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
 
@@
 In this paper, we describe the digital transcription of the full NES dataset with the goal of making the dataset openly accessible to the research community.
+Specifically, our objective was to exactly reproduce the contents of the original dataset rather than to evaluate the its integrity.
 We introduce and publish the data in an open format that requires no proprietary software.
 It can be easily downloaded, used for analysis, and amended.
-The provided summary statistics and figures also allow users to quickly assess the utility of the data..
+The provided summary statistics and figures also allow users to quickly assess the utility of the data.
 Finally, the code and raw data files are provided to facilitate the extraction of fields not represented in our completed dataset (mostly phytoplankton diversity data).
@@
 
-Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analysis.
+Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analyses.
 Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.
@@
 For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
-As such, NES data values represent the mean of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.
+As such, NES data values represent the mean of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than summer measurements taken in a single year.
 
@@
 The data for each NES region is contained in four separate files.
-We extracted the data from each file using automated techniques followed by manual quality assurance and checking.
+We extracted the data from each file using automated techniques followed by manual quality assurance and checking of each value.
 To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/).
@@
 Inaccurate OCR outputs were hand-corrected in the final database.
+Because our goal was to reproduce the data from the original reports and not to verify the technical correctness of the original data, we only changed values if they did not match the orginal data reports.
+For example, we did not change data from the five NES lakes that had phosphate ($PO^4$) values exceeding their corresponding total phosphorus ($TP$) values despite the fact that this is not physically possible ($PO^4$ is a component of $TP$).
 
 We provide the final dataset in an open non-proprietary format (comma-delimited, *.csv).
-We generated metadata descriptions from the contents of the original scanned reports.
+In addition, we generated metadata descriptions from the contents of the original scanned reports.
 All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts \citep{R-core}.
@@
 
-%\begin{table*}[t]
-%\caption{TEXT}
-%\begin{tabular}{column = lcr}
-%\tophline
-%
-%\middlehline
-%
-%\bottomhline
-%\end{tabular}
-%\belowtable{} % Table Footnotes
-%\end{table*}
-
-Although the overall number of lakes was similar among regions, they differed in the proportion of lakes classified as impoundments rather than natural lakes.
-For example, slightly more than half of the lakes studied (462 of 775) were classified as impoundments yet the northeastern region had only 54 impoundments and the southeastern region had 168 impoundments.
+In addition to differences in the total number of lakes measured in each region, there were also differences in the proportion of lakes classified as impoundments rather than as natural lakes.
+For example, slightly more than half of all the lakes studied (462 of 775) were classified as impoundments yet the northeastern region had only 54 impoundments while the southeastern region had 168 impoundments.
 Conversely, the number of natural lakes sampled in the northeastern region (146 lakes) was more than double that of any other region (77, 48, and 42, for southeastern, western, and central United States, respectively).

### 2017-12-04 08_manuscript/post_response.tex
typo fixes
fix figure legend colors

@@
   \includegraphics[width=12cm]{points_regions-crop.pdf}
-  \caption{Survey locations colored by sampling year (1972 northeastern-red, 1973 southeastern-green, 1974 central-blue, 1975 western-grey).}\label{fig:points_regions}
+  \caption{Survey locations colored by sampling year (1972 northeastern-light blue, 1973 southeastern-blue, 1974 central-light green, 1975 western-green).}\label{fig:points_regions}
 \end{figure*}
@@
 In this paper, we describe the digital transcription of the full NES dataset with the goal of making the dataset openly accessible to the research community.
-Specifically, our objective was to exactly reproduce the contents of the original dataset rather than to evaluate the its integrity.
+Specifically, our objective was to exactly reproduce the contents of the original dataset rather than to evaluate its scientific integrity.
 We introduce and publish the data in an open format that requires no proprietary software.
@@
 
-% \begin{acknowledgements}
-% TEXT
-% \end{acknowledgements}
+\begin{acknowledgements}
+
+\end{acknowledgements}
 

### 2017-12-04 08_manuscript/abstract.tex
add abstract make targets

--- 2017-11-07 08_manuscript/post_response.tex
@@
 \end{abstract}
-
-% \copyrightstatement{TEXT}
-
-\introduction  %% \introduction[modified heading if necessary]
-Effective management of inland freshwater lakes requires an understanding of the factors that affect water quality and how these factors change over time.
-One of these factors, termed eutrophication, occurs when excess nutrient inputs from human activities fuels increases in algal growth which can cause hypoxia and decreases in water clarity.
-Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture \citep{smith1999eutrophication, smith2014comment}.
-As human populations and their impacts continue to grow, eutrophication is expected to become more widespread \citep{bennett2001human, taranu2008quantifying}.
-Historical datasets are needed in order to track, understand, and manage eutrophication in lakes and reservoirs because they serve as an important baseline for modern studies.
-
-The U.S.
-Environmental Protection Agency (EPA) designed and implemented the National Eutrophication Survey (NES) in order to investigate the extent of eutrophication in freshwater lakes and reservoirs across the contiguous United States (US).
-Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.
-Each lake was sampled on a monthly basis for a period one year.
-Except for the phytoplankton distribution subset, which we did not transcribe \citep[see ][]{stomp2011large}, the NES data is provided as annual averages.
-Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}.
-Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure \ref{fig:points_regions}).
-In the remainder of the present paper we refer to the former two regions as simply the northeastern and southeastern regions.
-
-\begin{figure*}[t]
-  \includegraphics[width=12cm]{points_regions-crop.pdf}
-  \caption{Survey locations colored by sampling year (1972 northeastern-red, 1973 southeastern-green, 1974 central-blue, 1975 western-grey).}\label{fig:points_regions}
-\end{figure*}
-
-To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.
-For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance \citep{soballe1987large}.
-Also, it was used to predict eutrophication incidence in a Bayesian framework (Lamon and Stow 2004).
-Smaller portions of the data were used to explore drivers of nutrient loading \citep{stomp2011large,Brettreviewreassessmentlake2007}.
-Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is \citet{stomp2011large}, but their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.
-
-The present study is the first to leverage digital transcription technologies to unlock the full NES dataset.
-In this paper, we describe the digital transcription of the full NES dataset with the goal of making the dataset openly accessible to the research community.
-Specifically, our objective was to exactly reproduce the contents of the original dataset rather than to evaluate the its integrity.
-We introduce and publish the data in an open format that requires no proprietary software.
-It can be easily downloaded, used for analysis, and amended.
-The provided summary statistics and figures also allow users to quickly assess the utility of the data.
-Finally, the code and raw data files are provided to facilitate the extraction of fields not represented in our completed dataset (mostly phytoplankton diversity data).
-
-\section{Methods}
-
-Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analyses.
-Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.
-More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication \citep{nes-methods}.
-Due to historical nature of the dataset, the NES sampling design differs from more modern efforts \citep{nla-methods}.
-For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer.
-As such, NES data values represent the mean of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than summer measurements taken in a single year.
-
-We obtained the NES archival scanned reports from the EPA National Service Center for Environmental Publications (available at: https://www.epa.gov/nscep).
-The data for each NES region is contained in four separate files.
-We extracted the data from each file using automated techniques followed by manual quality assurance and checking of each value.
-To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/).
-Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) \citep{R-tesseract, smith2007overview}.
-The output of these initial extraction steps were recorded in a set of “raw data” files where each file contains the raw unprocessed text of each document page.
-The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package \citep{R-nesR}.
-Finally, all values in the database were manually checked for accuracy against the original scanned reports.
-Inaccurate OCR outputs were hand-corrected in the final database.
-Because our goal was to reproduce the data from the original reports and not to verify the technical correctness of the original data, we only changed values if they did not match the orginal data reports.
-For example, we did not change data from the five NES lakes that had phosphate ($PO^4$) values exceeding their corresponding total phosphorus ($TP$) values despite the fact that this is not physically possible ($PO^4$ is a component of $TP$).
-
-We provide the final dataset in an open non-proprietary format (comma-delimited, *.csv).
-In addition, we generated metadata descriptions from the contents of the original scanned reports.
-All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts \citep{R-core}.
-Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages \citep{R-plyr,R-sp}.
-
-\section{Results}
-
-The final NES dataset contains observations from 775 lakes and the distribution of these lakes was spatially variable.
-Although there were more lakes measured in the northeastern and southeastern United States, the number of locations was close to evenly distributed among the remaining regions (Figure \ref{fig:points_regions}, Table \ref{table:n}).
-Specifically, the number of lakes sampled in each region were as follows: northeastern - 200 lakes, southeastern - 245 lakes, central - 177 lakes, and western, 153 lakes.
-
-\begin{table*}
-\caption{Number of measurements (n) for each variable in each NES region.}\label{table:n}
-\begin{tabular}{lllll}
-\tophline
-Variable & Western & Central & Northeastern & Southeastern\\
-\middlehline
-Drainage area & 122 & 138 & 171 & 232\\
-
-Surface area & 152 & 177 & 200 & 245\\
-
-Mean depth & 149 & 174 & 174 & 242\\
-
-Total inflow & 124 & 138 & 170 & 232\\
-
-Retention time & 124 & 140 & 158 & 230\\
-
-Alkalinity & 153 & 177 & 200 & 245\\
-
-Conductivity & 153 & 176 & 200 & 245\\
-
-Secchi depth & 153 & 177 & 200 & 245\\
-
-Total P & 153 & 177 & 200 & 245\\
-
-Total inorg. P & 153 & 177 & 200 & 245\\
-
-Total inorg. N & 153 & 177 & 200 & 245\\
-
-Total N & 152 & 176 & 1 & 245\\
-
-P pt. source mun. & 52 & 83 & 139 & 189\\
-
-P pt. source ind. & 7 & 1 & 10 & 24\\
-
-P pt. source sep. & 65 & 88 & 111 & 175\\
-
-P nonpt. source & 122 & 133 & 167 & 231\\
-
-P total inputs & 122 & 133 & 167 & 231\\
-
-N pt. source mun. & 52 & 84 & 139 & 189\\
-
-N pt. source ind. & 7 & 1 & 8 & 22\\
-
-N pt. source sep. & 77 & 90 & 111 & 184\\
-
-N nonpt. source & 122 & 129 & 167 & 231\\
-
-N total inputs & 122 & 129 & 167 & 231\\
-
-P total exports & 119 & 132 & 167 & 227\\
-
-P retention & 99 & 115 & 144 & 201\\
-
-P load per area & 122 & 133 & 167 & 231\\
-
-N total exports & 119 & 133 & 166 & 227\\
-
-N retention & 88 & 111 & 122 & 170\\
-
-N load per area & 122 & 135 & 167 & 231\\
-\bottomhline
-\end{tabular}
-\end{table*}
-
-In addition to differences in the total number of lakes measured in each region, there were also differences in the proportion of lakes classified as impoundments rather than as natural lakes.
-For example, slightly more than half of all the lakes studied (462 of 775) were classified as impoundments yet the northeastern region had only 54 impoundments while the southeastern region had 168 impoundments.
-Conversely, the number of natural lakes sampled in the northeastern region (146 lakes) was more than double that of any other region (77, 48, and 42, for southeastern, western, and central United States, respectively).
-
-We observed substantial spatial variation in many of the individual lake characteristics.
-For example, lakes in the eastern sub-regions were generally smaller and shallower than lakes in the western sub-region (Table \ref{table:means}).
-In addition, lakes in the western sub-region generally had higher alkalinity and higher water clarity (Figure \ref{fig:alkalinity}, \ref{fig:secchi}).
-Lakes with particularly low alkalinity were found in coastal areas, whereas lakes with particularly high alkalinity were found in Nevada, western Washington, and parts of North Dakota.
-Comparisons among regions was easy for some well-sampled lake chemistry parameters such as total phosphorus but more difficult for undersampled lake chemistry parameters.
-A particularly extreme example of this difficulty was total nitrogen measurements in the eastern region, as this parameter was only measured for a single lake (Table \ref{table:n}).
-
-\begin{table*}[t]
-\caption{Mean and standard deviation (sd) for each variable in each NES region.}\label{table:means}
-\begin{tabular}{lllllllllllll}
-\tophline
-\multicolumn{1}{l}{Region} & \multicolumn{3}{c}{Western} & \multicolumn{3}{c}{Central} & \multicolumn{3}{c}{Northeastern} & \multicolumn{3}{c}{Southeastern} \\ \cline{1-1} \cline{2-4} \cline{5-7} \cline{8-10} \cline{11-13}
-Variable & Mean &   & sd & Mean &   & sd & Mean &   & sd & Mean &   & sd\\
-\middlehline
-Drainage area ($km^{2}$) & 2.5e+04 & ± & 7.8e+04 & 2.1e+04 & ± & 7.5e+04 & 3.2e+03 & ± & 1.4e+04 & 5.3e+03 & ± & 1.4e+04\\
-
-Surface area ($km^{2}$) & 44.57 & ± & 99.83 & 54.38 & ± & 1.4e+02 & 27.25 & ± & 99.01 & 42.7 & ± & 1.4e+02\\
-
-Mean depth ($m$) & 16.71 & ± & 27.08 & 5.97 & ± & 4.49 & 7 & ± & 9.37 & 6.4 & ± & 6.07\\
-
-Total inflow ($m^{3} \cdot s^{-1}$) & 52.1 & ± & 1.1e+02 & 31.82 & ± & 71.77 & 23.1 & ± & 65.26 & 82.6 & ± & 2.3e+02\\
-
-Retention time ($years$) & 7.27 & ± & 43.32 & 2.78 & ± & 6.98 & 2.01 & ± & 4.77 & 0.59 & ± & 1.12\\
-
-Alkalinity ($mg \cdot l^{-1}$) & 1.7e+02 & ± & 3.7e+02 & 1.5e+02 & ± & 91.51 & 1.2e+02 & ± & 1.6e+02 & 72.18 & ± & 66.25\\
-
-Conductivity ($uohm$) & 4.9e+02 & ± & 1.0e+03 & 6.4e+02 & ± & 7.6e+02 & 3.3e+02 & ± & 4.0e+02 & 2.5e+02 & ± & 2.2e+02\\
-
-Secchi depth ($m$) & 2.86 & ± & 2.64 & 1.2 & ± & 0.91 & 1.81 & ± & 1.71 & 1.22 & ± & 0.82\\
-
-Total P ($mg \cdot l^{-1}$) & 0.07 & ± & 0.13 & 0.11 & ± & 0.16 & 0.16 & ± & 0.35 & 0.12 & ± & 0.27\\
-
-Total inorg. P ($mg \cdot l^{-1}$) & 0.04 & ± & 0.11 & 0.04 & ± & 0.07 & 0.11 & ± & 0.3 & 0.05 & ± & 0.15\\
-
-Total inorg. N ($mg \cdot l^{-1}$) & 0.14 & ± & 0.23 & 0.33 & ± & 0.58 & 0.47 & ± & 0.66 & 0.72 & ± & 0.91\\
-
-Total N ($mg \cdot l^{-1}$) & 0.62 & ± & 0.65 & 1.22 & ± & 1.11 & 0.12 & ± & NA & 1.56 & ± & 1.25\\
-
-P pt. source mun. ($kg \cdot yr^{-1}$) & 2.5e+04 & ± & 8.7e+04 & 2.3e+04 & ± & 5.6e+04 & 3.5e+04 & ± & 1.5e+05 & 4.5e+04 & ± & 1.1e+05\\
-
-P pt. source ind. ($kg \cdot yr^{-1}$) & 2.5e+04 & ± & 4.0e+04 & 1.3e+04 & ± & NA & 2.7e+04 & ± & 4.9e+04 & 1.7e+04 & ± & 4.5e+04\\
-
-P pt. source sep. ($kg \cdot yr^{-1}$) & 56.62 & ± & 1.4e+02 & 60.62 & ± & 93.67 & 1.6e+02 & ± & 3.4e+02 & 98.55 & ± & 2.3e+02\\
-
-P nonpt. source ($kg \cdot yr^{-1}$) & 1.4e+05 & ± & 4.2e+05 & 1.8e+05 & ± & 6.8e+05 & 5.6e+04 & ± & 2.1e+05 & 1.9e+05 & ± & 5.5e+05\\
-
-P total inputs ($kg \cdot yr^{-1}$) & 1.5e+05 & ± & 4.7e+05 & 2.0e+05 & ± & 7.0e+05 & 8.7e+04 & ± & 3.4e+05 & 2.3e+05 & ± & 5.8e+05\\
-
-N pt. source mun. ($kg \cdot yr^{-1}$) & 7.8e+04 & ± & 2.5e+05 & 7.3e+04 & ± & 1.7e+05 & 1.4e+05 & ± & 5.4e+05 & 1.4e+05 & ± & 3.8e+05\\
-
-N pt. source ind. ($kg \cdot yr^{-1}$) & 2.3e+07 & ± & 6.1e+07 & 4.0e+03 & ± & NA & 1.6e+05 & ± & 4.2e+05 & 1.7e+05 & ± & 5.6e+05\\
-
-N pt. source sep. ($kg \cdot yr^{-1}$) & 5.7e+06 & ± & 5.0e+07 & 2.2e+03 & ± & 3.5e+03 & 4.3e+03 & ± & 5.5e+03 & 3.3e+03 & ± & 6.7e+03\\
-
-N nonpt. source ($kg \cdot yr^{-1}$) & 1.8e+06 & ± & 4.9e+06 & 1.8e+06 & ± & 4.4e+06 & 1.2e+06 & ± & 4.1e+06 & 3.1e+06 & ± & 8.9e+06\\
-
-N total inputs ($kg \cdot yr^{-1}$) & 6.8e+06 & ± & 5.7e+07 & 1.8e+06 & ± & 4.3e+06 & 1.3e+06 & ± & 4.6e+06 & 3.2e+06 & ± & 9.0e+06\\
-
-P total exports ($kg \cdot yr^{-1}$) & 6.2e+04 & ± & 1.7e+05 & 7.4e+04 & ± & 1.9e+05 & 7.3e+04 & ± & 3.1e+05 & 1.9e+05 & ± & 6.3e+05\\
-
-P retention (\%) & 47.77 & ± & 28.5 & 57.55 & ± & 26.01 & 36.93 & ± & 25.2 & 42.7 & ± & 23.34\\
-
-P load per area($g / m^{2} / day$) & 5.61 & ± & 21.36 & 3.3 & ± & 9.2 & 28.46 & ± & 97.49 & 9.43 & ± & 17.06\\
-
-N total exports ($kg \cdot yr^{-1}$) & 1.6e+06 & ± & 4.0e+06 & 1.2e+06 & ± & 2.8e+06 & 1.2e+06 & ± & 4.9e+06 & 3.0e+06 & ± & 8.3e+06\\
-
-N retention (\%) & 39.33 & ± & 27.13 & 43.41 & ± & 23.97 & 28.41 & ± & 23.62 & 26.28 & ± & 18.85\\
-
-N load per area ($g / m^{2} / day$) & 1.8e+02 & ± & 1.1e+03 & 42.67 & ± & 1.1e+02 & 2.8e+02 & ± & 9.1e+02 & 1.3e+02 & ± & 2.4e+02\\
-\bottomhline
-\end{tabular}
-\end{table*}
-
-\begin{figure*}[t]
-  \includegraphics[width=12cm]{alkalinity-crop.pdf}
-  \caption{Map of log-scaled alkalinity (mg/L) interpolated using inverse distance weighting.}\label{fig:alkalinity}
-\end{figure*}
-
-\begin{figure*}[t]
-  \includegraphics[width=12cm]{secchi-crop.pdf}
-  \caption{Map of secchi depth (m) interpolated using inverse distance weighting.}\label{fig:secchi}
-\end{figure*}
-
-The ability to examine these spatial trends was made possible by our optical character recognition procedure which had 6 - 17\% accuracy depending on region and archival report scan quality.
-In total, we carried out approximately 5,000 hand-corrections to the automated data product as part of our manual quality control review.
-A total of approximately 650 lakes had values for at least 80\% of the total number of variables shown in Table \ref{table:n}.
-On an individual lake basis, the most common “missing” data was nutrient loading estimates for individual point and nonpoint-source components.
-In many cases, this data may not actually be missing but it may have not been a component of the budget for that particular lake.
-For example, not all lakes have industrial land use so no data is expected in these cases.
-
-\section{Discussion}
-
-We have demonstrated an approach for rescuing historical data from scanned documents.
-In particular, our approach involved a two-step process of automated data scraping followed by hand-curation and quality assurance.
-Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records \citep{drinkwater2014use}.
-Unfortunately, optical character recognition technology does not have absolute accuracy.
-In our case, transcription was hampered by poor print and scan quality of the source paper documents.
-We discovered through our manual validation procedure that the OCR computations produced inaccurate values in approximately 6 - 17\% of the cells in the complete dataset (n = 4836).
-We expect that accuracy could be improved by experimenting with varying the window size of the local adaptive thresholding algorithm relative to the document font size.
-Our ability to experiment with thresholding window size was limited due to the computationally expensive nature of these extractions.
-
-The end result of our approach was data from every lake and nearly every variable in the NES survey dataset.
-The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by \citet{stomp2011large}.
-The results of the present study could be used to explore anthropogenic and environmental drivers of lake eutrophication as well as to verify previously documented trends.
-One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes \citep{nla-methods}.
-This reanalysis considered population level trends in the NES lakes but did not consider trends in individual lakes or potential environmental drivers contributing to observed trends.
-On a population basis, the NLA reanalysis found that less than 30\% of the NES lakes had increased chlorophyll and phosphorus concentrations.
-The results of the present study could be used to verify these claims as well as to compare the NES data with more recent work such as the 2012 National Lakes Assessment.
-Note that sampling techniques may differ from current techniques, so care should be given when making comparisons.
-In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time.
-Such data is critical to a variety of hydrological and water quality modelling efforts \citep{Brettreviewreassessmentlake2007}.
-
-Although our goal was to digitally transcribe the full NES dataset to facilitate studies on historical nutrient loading, it is worth noting the similarities between the present study and other scientific record digitization initiatives.
-Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences \citep{allan2011international, freeman2017icoads}.
-To our knowledge, the present study is the first large-scale attempt at digitization of historical limnology records.
-We hope that by making our analysis open and reproducible we will inspire future efforts at recovering important records from the pre-digital era.
-
-%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
-%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
-
-% \codeavailability{TEXT} %% use this section when having only software code available
-
-% \dataavailability{} %% use this section when having only data sets available
-
-\codedataavailability{
-Original scanned reports from the EPA are available from the EPA National Service Center for Environmental Publications (https://www.epa.gov/nscep).
-Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z.
-The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions \citep{nes}.
-Moreover, the R package and R code used to scrape and analyze the data are provided by \citet{R-nesR} so that the methods may be reproduced and openly available for (re)use.
-All figures and summary statistics were generated by R scripts available in the data supplement linked above.
-} %% use this section when having data sets and software code available
-
-% \appendix
-% \section{}    %% Appendix A
-% 
-% \subsection{}     %% Appendix A1, A2, etc.
-
-% \noappendix       %% use this to mark the end of the appendix section
-
-%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
-
-%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
-%% They will be correctly named automatically.
-
-%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
-%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:
-
-\appendixfigures  %% needs to be added in front of appendix figures
-
-\appendixtables   %% needs to be added in front of appendix tables
-
-%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.
-
-\authorcontribution{
-All authors contributed to data quality assurance and edited the article text.
-JS conceived the study and implemented the optical character recognition code.
-CF, DK, and RN performed the data analysis and made figures.
-KK, HM, and JS wrote major parts of the manuscript text.
-} %% optional section
-
-\competinginterests{
-The authors declare that they have no conflict of interest.
-} %% this section is mandatory even if you declare that no competing interests are present
 

### 2017-12-04 08_manuscript/post_response.tex
add acknowledgements

@@
 \begin{acknowledgements}
-
+  This work was developed as part of the Reproducible Quantitative Methods course (https://cbahlai.github.io/rqm-template/) led by Dr.
+Christie Bahlai which was funded by the Mozilla Foundation, the Leona M.
+and Harry B.
+Helmsley Charitable Trust, the Michigan State University Program in Ecology and Evolutionary Biology, the BEACON Center for Evolution in Action, and the Kellogg Biological Station Long Term Ecological Research Site (NSF-DEB \#1027253).
 \end{acknowledgements}

### 2017-12-04 08_manuscript/post_response.tex
update doi

@@
 The goal of our approach was to strike a balance between efficiency and data quality by combining hand-entry of data with digital transcription technologies.
-The finished database contains information on the physical characteristics, hydrology, and water quality of about 800 lakes in the contiguous United States (doi:10.5063/F10G3H3Z).
+The finished database contains information on the physical characteristics, hydrology, and water quality of about 800 lakes in the contiguous United States \citep{nes}.
 Ultimately, this database could be combined with more recent studies to generate meta-analyses of water quality trends and spatial variation across the continental United States.
@@
 Original scanned reports from the EPA are available from the EPA National Service Center for Environmental Publications (https://www.epa.gov/nscep).
-Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z.
+Our cleaned and useable data are available for download at \citet{nes}.
 The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions \citep{nes}.

### 2019-12-06 README.Rmd
try adding doi badge
try adding doi badge 2
fix image link, rebuild
add workflow description and contributing instructions

--- 2017-07-29 README.Rmd
@@
 
+[![DOI](https://img.shields.io/badge/DOI-10.5063/F16D5R2B-blue.svg)](https://doi.org/10.5063/F16D5R2B)
+
 This respository contains a digitally transcribed copy ([nes_data.csv](https://github.com/ReproducibleQM/NES/raw/master/nes_data.csv)) of the National Eutrophication Survey (NES) dataset.
@@
 points <- image_read("06_images/points.jpeg")
-image_write(image_trim(points), "06_images/points_trim.jpeg")
+image_write(image_trim(points), 
+            "06_images/points_trim.jpeg")
 ```
@@
 
+## Worflow
+
+1. Use the [nesR](https://github.com/jsta/nesR) package to generate the files in [/02_raw_data](/02_raw_data).
+
+2. Hand check the files in [/02_raw_data/merged_data](/02_raw_data/merged_data) against the pdfs in [/01_archival_pdfs](/01_archival_pdfs).
+
+3. Combine files with [05_analysis_scripts/PDF_Merge.R](05_analysis_scripts/PDF_Merge.R).
+
+## Contributing
+
+We've combed the data to try and find all the transciption errors but it's difficult to catch them all.
+If you find any errors please open an [issue](https://github.com/ReproducibleQM/NES/issues/) or submit a pull request against the files in [/03_qa_data](/03_qa_data).
+
 ## References

### 2019-12-09 05_analysis_scripts/add_chl.R
automate nes_data.csv building

@@ -1,19 +1,19 @@
 # Prep for QA ####
-raw_paths <- c("../nesR/474/res.csv",
-               "../nesR/475/res.csv",
-               "../nesR/476/res.csv",
-               "../nesR/477/res.csv")
-surveys <- 474:477
-
-dt_raw <- lapply(seq_len(length(surveys)), function(x){
-  dt <- read.csv(raw_paths[x], stringsAsFactors = FALSE)
-  dt$survey <- surveys[x]
-  dt
-})
-
-dt <- do.call("rbind", dt_raw)
-dt <- dt[, c("survey", "pagenum", "chl")]
-write.csv(dt, "02_raw_data/merged_data/chl.csv", row.names = FALSE)
+# raw_paths <- c("../nesR/474/res.csv",
+#                "../nesR/475/res.csv",
+#                "../nesR/476/res.csv",
+#                "../nesR/477/res.csv")
+# surveys <- 474:477
+# 
+# dt_raw <- lapply(seq_len(length(surveys)), function(x){
+#   dt <- read.csv(raw_paths[x], stringsAsFactors = FALSE)
+#   dt$survey <- surveys[x]
+#   dt
+# })
+# 
+# dt <- do.call("rbind", dt_raw)
+# dt <- dt[, c("survey", "pagenum", "chl")]
+# write.csv(dt, "02_raw_data/merged_data/chl.csv", row.names = FALSE)
 
 # after QA ####
 

### 2019-12-09 README.Rmd
update essd ref
typo fix
update doi

@@
 
-[![DOI](https://img.shields.io/badge/DOI-10.5063/F16D5R2B-blue.svg)](https://doi.org/10.5063/F16D5R2B)
+[![DOI](https://img.shields.io/badge/DOI-10.5063/F1CZ35HF-blue.svg)](https://doi.org/10.5063/F1CZ35HF)
 
@@
 
-## Worflow
+## Workflow
 
@@
 
-Stachelek, J., Ford, C., Kincaid, D., King, K., Miller, H., Nagelkirk, R. (_in prep_) The National Eutrophication Survey: lake characteristics and historical nutrient concentrations.
+Stachelek, J., Ford, C., Kincaid, D., King, K., Miller, H.
+and Nagelkirk, R., 2018.
+The National Eutrophication Survey: lake characteristics and historical nutrient concentrations.
+Earth System Science Data, 10(1), pp.81-86.
+(https://doi.org/10.5194/essd-10-81-2018)
 
